---
title: "Demonstratios of particular regression and ANOVA concepts"
subtitle: "From simple vs multiple regression to Type I/II/III ANOVA tests and handling interactions"
format:
  html:
    embed-resources: true
    code-fold: false
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

```{r}
#| include: false
set.seed(201)

# Core libraries
library(tibble)
library(ggplot2)
library(cowplot)
library(car)       # for Type II / III
library(dplyr)
library(broom)

# Theme
theme_cowplot_bg <- function(font_size = 14) {
  cowplot::theme_cowplot(font_size = font_size) +
    theme(
      plot.background = element_rect(fill = "white"),
      panel.background = element_rect(fill = "white")
    )
}

knitr::opts_chunk$set(echo = TRUE)
```

# Where are we?

* Described the general linear model and derived estimators and their distributions:

  * $\hat{\mathbf{\beta}} = (\symbf{X}^\top \symbf{X})^{-1}\symbf{X}^\top \symbf{Y}$ with
    $\hat{\mathbf{\beta}} \sim \mathcal N\left(\mathbf{\beta},\sigma^2(\symbf{X}^\top\symbf{X})^{-1}\right)$.
  * Unbiased variance estimator:
    $s^2=\hat{\sigma}^2_{\text{unbiased}} = \dfrac{1}{n-k}\sum_{i=1}^n (y_i-\hat y_i)^2$
    with $\dfrac{(n-k)s^2}{\sigma^2} \sim \chi^2_{n-k}$.

# Where are we going?

- Comparing estimators between simple and multiple regression, showing their numerical differences and explaining why they differ.
- Comparing different ANOVA types (Type I, Type II, Type III) and their relations with regression.
- Providing a final recommendation for handling testing for effects in the presence of interactions in multiple regression.

# Set-up code 

```{r}
#| eval: false
library(tibble)
library(ggplot2)
library(cowplot)
library(car)       # for Type II / III
library(dplyr)
library(broom)
theme_cowplot_bg <- function(font_size = 14) {
  cowplot::theme_cowplot(font_size = font_size) +
    theme(
      plot.background = element_rect(fill = "white"),
      panel.background = element_rect(fill = "white")
    )
}
```

# Example 1 — Simulated advertising data (n = 100)

We simulate **radio** and **TV** spend and a revenue outcome. TV spend is partly driven by radio (collinearity), and revenue depends strongly on TV.

```{r}
# Advertising data
set.seed(201)
n_ad <- 100L
spend_vec_radio <- runif(n_ad, 0, 3)
spend_vec_tv <- 1.2 * spend_vec_radio + runif(n_ad, 0, 3)
revenue_vec <- 2 * spend_vec_tv + rnorm(n_ad, mean = 10, sd = 2)

ad_tbl <- tibble(
  spend_radio = spend_vec_radio,
  spend_tv = spend_vec_tv,
  revenue = revenue_vec
)

head(ad_tbl, 6)
```

## Simple regressions: revenue vs each predictor

### Correspondence between ANOVA and regression

First, we fit two simple linear regression models:

```{r}
m_tv_simple   <- lm(revenue ~ spend_tv,   data = ad_tbl)
m_radio_simple<- lm(revenue ~ spend_radio,data = ad_tbl)
```

Here we display the fitted values along with their confidence bands:

```{r}
# Plots
p_tv <- ggplot(ad_tbl, aes(spend_tv, revenue)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "TV spend", y = "Revenue") +
  theme_cowplot_bg()

p_radio <- ggplot(ad_tbl, aes(spend_radio, revenue)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Radio spend", y = "Revenue") +
  theme_cowplot_bg()

plot_grid(p_tv, p_radio, ncol = 2)
```

## Multiple regression: revenue \~ tv + radio (order does **not** change the fit)

```{r}
m_both_tvfirst   <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
m_both_radiofirst<- lm(revenue ~ spend_radio + spend_tv, data = ad_tbl)

# Coefficients identical (fits are the same)
coef(m_both_tvfirst)
coef(m_both_radiofirst)
all.equal(fitted(m_both_tvfirst), fitted(m_both_radiofirst))
```

### Why?

Regress each predictor on the other, take residuals, then regress revenue residuals on predictor residuals — this reproduces the **partial** effects in the multiple regression.

```{r}
# Regress radio on TV; TV on radio
r_radio_given_tv <- resid(lm(spend_radio ~ spend_tv, data = ad_tbl))
r_tv_given_radio <- resid(lm(spend_tv ~ spend_radio, data = ad_tbl))

# Residualise revenue on the "other" predictor
r_y_given_tv    <- resid(lm(revenue ~ spend_tv,    data = ad_tbl))
r_y_given_radio <- resid(lm(revenue ~ spend_radio, data = ad_tbl))

# Residual-on-residual slopes
fwl_radio <- lm(r_y_given_tv    ~ r_radio_given_tv)
fwl_tv    <- lm(r_y_given_radio ~ r_tv_given_radio)

coef(fwl_radio) |> signif(2)  # intercept ~ 0; slope = partial effect of radio
coef(fwl_tv) |> signif(2)    # intercept ~ 0; slope = partial effect of TV

# Compare to the multiple regression slopes:
coef(m_both_tvfirst) |> signif(2)
```

We need not residualise the response variable (essentially because whatever the TV would have predicted, the residuals of radio could not predict):

```{r}
coef(lm(revenue ~ r_radio_given_tv, data = ad_tbl)) |>
  signif(2)  # radio partial slope
coef(lm(revenue ~ r_tv_given_radio, data = ad_tbl)) |>
  signif(2)  # TV
```

This goes towards explaining why the standard errors increase for correlated predictors.
In multiple regression, each explanatory variable is in a sense not modelled as is, but rather only whatever could not be explained by the other variables.
When the other variables predict it well, there is little left over, which means that the explanatory variable has little variability left.
For orthogonal predictors (which we now have), we have that $\mathrm{Var}(\hat{\beta_i}) = \frac{\sigma^2}{n\mathrm{Var}(X_i)}$, which increases with decreasing variability of $X_i$.

<!--

### QR decomposition view (orthogonal projections)

QR (with `qr()`) shows that OLS computes $\hat\beta$ using orthogonalised columns of $X$; hence permutation of columns does not change the fitted values.

```{r}
#| eval: false
X <- model.matrix(m_both_tvfirst)   # [Intercept, spend_tv, spend_radio]
y <- ad_tbl$revenue

qrX <- qr(X)
Q <- qr.Q(qrX)       # orthonormal basis
R <- qr.R(qrX)       # upper triangular

beta_qr <- backsolve(R, t(Q) %*% y)
cbind(beta_via_qr = as.vector(beta_qr),
      beta_lm     = coef(m_both_tvfirst))
max_abs_diff <- max(abs(beta_qr - coef(m_both_tvfirst)))
max_abs_diff
```

!-->

## ANOVA for simple linear regression

For a single predictor, `anova(lm(y ~ x))` matches the $F$-test on the slope (which has the same p-value as the $t$-test on the slope, but the square of the $t$-statistic).

```{r}
anova(lm(revenue ~ spend_tv, data = ad_tbl))   # TV
anova(lm(revenue ~ spend_radio, data = ad_tbl)) # Radio
```

```{r}
summary(m_tv_simple)
summary(m_radio_simple)
```

## Type I (sequential) sums of squares **depend on order**

The default `anova()` on a multiple regression is **Type I** (sequential). With correlated predictors, the first-entered variable can appear much more (or less) significant.

```{r}
anova(m_both_tvfirst)    # TV entered first
anova(m_both_radiofirst) # Radio entered first

# Note the second line’s SS and p-value change with order.
```

Importantly, the variable that entered last has exactly the same test results (p-value) as that from the multiple linear regression including both variables:

```{r}
lm(revenue ~ spend_radio + spend_tv, data = ad_tbl) |>
  summary()
```

So, regression always treats each variable as if they entered last.

---

# Example 2 — Full vs reduced model for a multi-level factor (global F test)

We now **do not** reuse the advertising data. Simulate a 3-level factor `region` with different means; show that:

1. `anova(full, reduced)` dropping the **entire** factor is a partial $F$ test.

2. The same result arises from the hand-coded quadratic form

$$
F=\frac{(L\hat\beta)^\top\left\{L\widehat{\text{Cov}}(\hat\beta)L^\top\right\}^{-1}(L\hat\beta)}{q}
$$

with $q=\text{rank}(L)$.

```{r}
set.seed(42)
n_region <- 300L
region <- factor(sample(c("North","Central","South"), size = n_region, replace = TRUE))
# True means by region; add noise
mu <- c(North = 12, Central = 10, South = 14)
y_region <- mu[region] + rnorm(n_region, sd = 2.5)

reg_tbl <- tibble(y = as.numeric(y_region), region)

# Full: with region; Reduced: intercept only
m_full    <- lm(y ~ region, data = reg_tbl)
m_reduced <- lm(y ~ 1,      data = reg_tbl)

anova(m_reduced, m_full)  # partial F for the factor as a whole
summary(m_full)
```

### Hand-coded $F$ — match the partial F above

```{r}
# Model matrix and estimated coefficients / covariance
X_full <- model.matrix(m_full)                   # columns: (Intercept), regionNorth, regionSouth
beta_hat <- coef(m_full)
sigma2_hat <- sum(resid(m_full)^2) / df.residual(m_full)
XtX_inv <- solve(t(X_full) %*% X_full)
Cov_beta <- sigma2_hat * XtX_inv

# Build L to jointly test all region dummies = 0 (i.e., drop entire factor)
cols <- colnames(X_full)
idx <- which(cols %in% c("regionNorth","regionSouth"))
L <- diag(length(beta_hat))[idx, , drop = FALSE]  # selects those coefficients

q <- qr(L)$rank
num <- t(L %*% beta_hat) %*% solve(L %*% Cov_beta %*% t(L)) %*% (L %*% beta_hat)
F_hand <- as.numeric(num / q)

# Compare to partial F from anova
F_aov <- anova(m_reduced, m_full)$`F`[2]
data.frame(F_hand = F_hand, F_from_anova = F_aov)
```

---

# Example 3 — Interaction model: drug × sex (n = 50)

We encode `sex` as 0 = female, 1 = male; `drug` as 0/1. We simulate a **real interaction** so simple Type III vs Type II decisions matter.

```{r}
set.seed(777)
n_int <- 100L
sex  <- rbinom(n_int, 1, 0.5)             # 0=female, 1=male
drug <- rbinom(n_int, 1, 0.5)

# True model: baseline 10, drug effect +2 in females, ~0 in males (interaction -2),
# sex main effect +1; noise sd=2
y <- 10 + 1*sex + 2*drug + (-2)*(sex*drug) + rnorm(n_int, sd = 2)

int_tbl <- tibble(y, sex = factor(sex, levels = c(0,1), labels = c("Female","Male")),
                  drug = factor(drug, levels = c(0,1), labels = c("No","Yes")),
                  sex_num = as.numeric(sex == "Male"),
                  drug_num = as.numeric(drug == "Yes"))

# Fit with interaction
m_int_full <- lm(y ~ sex * drug, data = int_tbl)
summary(m_int_full)
```

```{r}
# Visualise interaction
ggplot(int_tbl, aes(drug, y, colour = sex, group = sex)) +
  stat_summary(fun = mean, geom = "point", position = position_dodge(width = 0.2)) +
  stat_summary(fun = mean, geom = "line", position = position_dodge(width = 0.2)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1,
               position = position_dodge(width = 0.2)) +
  labs(title = "Drug × Sex interaction (means ± SE)",
       x = "Drug", y = "Outcome") +
  theme_cowplot_bg()
```

## Type II vs Type III (using `car::Anova`)

* **Type II** (hierarchical): tests each **main effect** *after* the other main effect, **but without** the interaction. Sensible if the interaction is absent (or negligible), and respects hierarchy.
* **Type III**: tests each term **with** all other terms in the model (including the interaction). For a binary `sex` coded with `Female = 0`, the Type III **drug** test is effectively “drug effect at **Female**”.

```{r}
Anova(m_int_full, type = 2)  # Type II
Anova(m_int_full, type = 3)  # Type III
```

### Testing the interaction **first**

A principled approach is to test the interaction via a partial $F$: compare the full model to a **main-effects-only** model.

```{r}
m_int_main <- lm(y ~ sex + drug, data = int_tbl)
anova(m_int_main, m_int_full)  # test of the interaction term
```

* If the interaction is **not** significant, one workflow is to drop it and use **Type II** to test main effects.
  *Caveat*: this is **data-driven model selection**, which can inflate power and does **not** preserve exact Type I error.
* Alternatively, **keep** the interaction and use **Type III** to ask context-specific questions (e.g., “effect for females”, if they are coded 0). This can be less powerful if weak, same-direction effects exist in both groups, because we do not pool.

### Simultaneous “any drug effect” test (main effect **and** interaction)

We can test **both** the drug main effect and its interaction jointly with an $F$ test:

$$
H_0: \beta_{\text{drug}} = 0 \ \text{and}\  \beta_{\text{sex:drug}} = 0.
$$

This asks whether the drug affects **either** sex.

```{r}
# Build L that extracts the drug main effect and drug:sex interaction(s)
X_int   <- model.matrix(m_int_full)
b_int   <- coef(m_int_full)
s2_int  <- sum(resid(m_int_full)^2) / df.residual(m_int_full)
Vb_int  <- s2_int * solve(t(X_int) %*% X_int)

# Identify columns
colnames(X_int)
# Terms are typically: (Intercept), sexMale, drugYes, sexMale:drugYes
L_any_drug <- rbind(
  "drugYes"           = as.numeric(names(b_int) == "drugYes"),
  "sexMale:drugYes"   = as.numeric(names(b_int) == "sexMale:drugYes")
)

q <- qr(L_any_drug)$rank
F_num <- t(L_any_drug %*% b_int) %*% solve(L_any_drug %*% Vb_int %*% t(L_any_drug)) %*% (L_any_drug %*% b_int)
F_any_drug <- as.numeric(F_num / q)
p_any_drug <- pf(F_any_drug, df1 = q, df2 = df.residual(m_int_full), lower.tail = FALSE)

data.frame(F_any_drug = F_any_drug, df1 = q, df2 = df.residual(m_int_full), p_value = p_any_drug)
```

Equivalently, one can use the `anova` function with the full and reduced model (not including any terms involving drug):

```{r}
m_no_drug <- lm(y ~ sex)
anova(m_no_drug, m_int_full)
```

