---
title: Motivating multiple linear regression
subtitle: Extension to SlidesRegW2S1GLM
format:
  pdf:
    embed-resources: true
---

```{r}
#| include: false
library(ggplot2)
library(tibble)
theme_cowplot_bg <- function() {
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )
}
```

# Introduction

Suppose that we wish to uncover the relationship between revenue and advertising spend on TV and radio.

Let's further assume that (known to us, as we're constructing this scenario) that radio spend has no effect on revenue, and that TV spend has a linear relationship with revenue.

# Scenarios

We'll consider two scenarios: one where companies only spend money on radio, and one where companies spend money on both TV and radio.

## Scenario 1

Suppose we have companies that spend only money on radio, and no money on TV.

Then, for say 100 observations, we might have the following data:

```{r}
set.seed(1811)
spend_vec_radio <- runif(1e2, 0, 3)
revenue_vec <- rnorm(1e2, mean = 10, sd = 2)
```

Note that above the simulated revenues are not dependent on the radio spend.

```{r}
p <- ggplot(
  data = data.frame(spend_vec_radio, revenue_vec),
  aes(x = spend_vec_radio, y = revenue_vec)
) +
  theme_cowplot_bg() + 
  cowplot::background_grid(major = "xy") +
  geom_point(col = "dodgerblue") +
  labs(x = "Radio spend", y = "Revenue")
p
```

If we add a line of best fit, we can see that it's basically flat:

```{r}
p +
  geom_smooth(method = "lm", se = FALSE, col = "orange", formula = y ~ x)
```

It's not perfectly flat just because of random noise in the data.

If we fit a linear regression to the data, we could see that there's no effect:

```{r}
lm(revenue_vec ~ spend_vec_radio) |> summary()
```

## Scenario 2

Now suppose that we have companies that spend money on TV, and that the more they spend on TV, they more they spend on radio:

```{r}
set.seed(0201)
spend_vec_radio <- runif(1e2, 0, 3)
spend_vec_tv <- spend_vec_radio * 1.2 + runif(1e2, 0, 3)
revenue_vec <- spend_vec_tv * 2 + rnorm(1e2, mean = 10, sd = 2)
```

Note that above the simulated revenues are dependent on the TV spend, and that the TV spend is related to the radio spend - but there is no direct relationship between radio spend and revenue (i.e. `spend_vec_radio` does not appear in the formula for `revenue_vec`).
So the effect of radio on revenue is the same as in the previous scenario.

If we plot the data, we can see that there's a relationshp between TV spend and revenue:

```{r}
ggplot(
  data = data.frame(spend_vec_tv, revenue_vec),
  aes(x = spend_vec_tv, y = revenue_vec)
) +
  theme_cowplot_bg() + 
  cowplot::background_grid(major = "xy") +
  geom_point(col = "dodgerblue") +
  labs(x = "TV spend", y = "Revenue") +
  geom_smooth(method = "lm", se = FALSE, col = "orange", formula = y ~ x)
```

The simple linear regression model would show that there's a relationship between TV spend and revenue:

```{r}
lm(revenue_vec ~ spend_vec_tv) |> summary()
```

Now, we also now find that there's a relationship between radio spend and revenue:

```{r}
ggplot(
  data = data.frame(spend_vec_radio, revenue_vec),
  aes(x = spend_vec_radio, y = revenue_vec)
) +
  theme_cowplot_bg() + 
  cowplot::background_grid(major = "xy") +
  geom_point(col = "dodgerblue") +
  labs(x = "Radio spend", y = "Revenue") +
  geom_smooth(method = "lm", se = FALSE, col = "orange", formula = y ~ x)
```

```{r}
lm(revenue_vec ~ spend_vec_radio) |> summary()
```

This is because radio spend is correlated with TV spend:

```{r}
ggplot(
  data = data.frame(spend_vec_tv, spend_vec_radio),
  aes(x = spend_vec_tv, y = spend_vec_radio)
) +
  theme_cowplot_bg() + 
  cowplot::background_grid(major = "xy") +
  geom_point(col = "dodgerblue") +
  labs(x = "TV spend", y = "Radio spend") +
  geom_smooth(method = "lm", se = FALSE, col = "orange", formula = y ~ x)
```

Now, if we fit a multiple linear regression of revenue against TV spend and radio spend, we can see that the model correctly identifies the relationship between TV and radio spend and revenue:

```{r}
lm(revenue_vec ~ spend_vec_tv + spend_vec_radio) |> summary()
```

There's a small negative effect of radio spend on TV estimated, but it's not statistically sigificant.