---
title: "Regression Week V: Prediciton"
author: "Miguel Rodo"
institute: "University of Cape Town"
aspectratio: 169
theme: Boadilla
format:
  beamer:
    embed-resources: true
    colorlinks: true
    include-in-header:
      file: preamble.tex
    classoption: "aspectratio=169"
---

```{r}
#| include: false
library(tibble)
library(ggplot2)
theme_cowplot_bg <- function(font_size = 14) {
  cowplot::theme_cowplot(font_size = font_size) +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )
}
knitr::opts_chunk$set(echo = TRUE)
```

# Predicting Individual Values

- Suppose we want to predict the range of individual future observations.
- We fit a linear regression model to our data.
- Then, we can obtain **prediction intervals** for new observations:
  - The prediction interval is an interval that contains the new observation with a certain probability.

# Confidence Intervals vs. Prediction Intervals

- **Confidence intervals** estimate the mean response at a given predictor value.
  - Source of variability: Uncertainty in estimating $\beta$.
- **Prediction intervals** estimate the range of individual future observations.
  - Sources of variability: Uncertainty in estimating $\beta$ and random variation in future observations.

# Example (ten observations)

```{r}
#| echo: false
n <- 10
x_vec <- runif(n, 0, 10)
y_vec <- 2 * x_vec + rnorm(n, 0, 2)
example_tbl <- tibble(x = x_vec, y = y_vec)
fit_vec <- lm(y ~ x, data = example_tbl)
x_vec_seq <- seq(min(x_vec), max(x_vec), length.out = 1e3)
y_vec_seq_pred <- predict(fit_vec, newdata = tibble(x = x_vec_seq), interval = "prediction")
y_vec_seq_conf <- predict(fit_vec, newdata = tibble(x = x_vec_seq), interval = "confidence")
plot_tbl_int <- tibble(
  x = rep(x_vec_seq, 2),
  lower = c(y_vec_seq_pred[, 2], y_vec_seq_conf[, 2]),
  upper = c(y_vec_seq_pred[, 3], y_vec_seq_conf[, 3]),
  interval_type = rep(c("Prediction", "Confidence"), each = length(x_vec_seq))
)

p <- ggplot(example_tbl, aes(x = x, y = y)) +
  theme_cowplot_bg() +
  cowplot::background_grid(major = "xy", minor = "none") +
  geom_point(col = "dodgerblue", alpha = 0.8) +
  geom_smooth(col = "orange", method = "lm", se = FALSE, size = 1.5) +
  geom_ribbon(
    data = plot_tbl_int |>
      dplyr::filter(interval_type == "Confidence"),
     aes(x = x, ymin = lower, ymax = upper, fill = interval_type),
     alpha = 0.5,
    inherit.aes = FALSE) +
  geom_line(
    data = plot_tbl_int |>
      dplyr::filter(interval_type == "Prediction"),
    aes(x = x, y = lower, col = interval_type),
    linetype = "11",
    size = 1
  ) +
  geom_line(
    data = plot_tbl_int |>
      dplyr::filter(interval_type == "Prediction"),
    aes(x = x, y = upper, col = interval_type),
    linetype = "11",
    size = 1
  ) +
  scale_colour_manual(values = c("Prediction" = "#e78ac3", "Confidence" = "#66c2a5")) +
  scale_fill_manual(values = c("Prediction" = "#e78ac3", "Confidence" = "#66c2a5")) +
  labs(
    x = "Predictor",
    y = "Response"
  ) +
  theme(
    legend.title = element_blank()
  )
path_fig <- projr::projr_path_get(
  "cache", "fig", "w5", "ci_vs_pred.pdf"
  )
ggsave(path_fig, plot = p, width = 14, height = 8, units = "cm")
```

\begin{center}
\includegraphics[width=0.8\linewidth  ]{`r path_fig`}
\end{center}

# Confidence intervals shrink to zero as sample size increases, but prediction intervals do not (100 observations)

```{r}
#| echo: false
n <- 200
x_vec <- runif(n, 0, 10)
y_vec <- 2 * x_vec + rnorm(n, 0, 2)
example_tbl <- tibble(x = x_vec, y = y_vec)
fit_vec <- lm(y ~ x, data = example_tbl)
x_vec_seq <- seq(min(x_vec), max(x_vec), length.out = 1e3)
y_vec_seq_pred <- predict(fit_vec, newdata = tibble(x = x_vec_seq), interval = "prediction")
y_vec_seq_conf <- predict(fit_vec, newdata = tibble(x = x_vec_seq), interval = "confidence")
plot_tbl_int <- tibble(
  x = rep(x_vec_seq, 2),
  lower = c(y_vec_seq_pred[, 2], y_vec_seq_conf[, 2]),
  upper = c(y_vec_seq_pred[, 3], y_vec_seq_conf[, 3]),
  interval_type = rep(c("Prediction", "Confidence"), each = length(x_vec_seq))
)

p <- ggplot(example_tbl, aes(x = x, y = y)) +
  theme_cowplot_bg() +
  cowplot::background_grid(major = "xy", minor = "none") +
  geom_point(col = "dodgerblue", alpha = 0.1) +
  geom_smooth(col = "orange", method = "lm", se = FALSE, size = 0.01) +
  geom_ribbon(
    data = plot_tbl_int |>
      dplyr::filter(interval_type == "Confidence"),
     aes(x = x, ymin = lower, ymax = upper, fill = interval_type),
     alpha = 0.5,
    inherit.aes = FALSE) +
  geom_line(
    data = plot_tbl_int |>
      dplyr::filter(interval_type == "Prediction"),
    aes(x = x, y = lower, col = interval_type),
    linetype = "11",
    size = 1
  ) +
  geom_line(
    data = plot_tbl_int |>
      dplyr::filter(interval_type == "Prediction"),
    aes(x = x, y = upper, col = interval_type),
    linetype = "11",
    size = 1
  ) +
  scale_colour_manual(values = c("Prediction" = "#e78ac3", "Confidence" = "#66c2a5")) +
  scale_fill_manual(values = c("Prediction" = "#e78ac3", "Confidence" = "#66c2a5")) +
  labs(
    x = "Predictor",
    y = "Response"
  ) +
  theme(
    legend.title = element_blank()
  )
path_fig <- projr::projr_path_get(
  "cache", "fig", "w5", "ci_vs_pred_n.pdf"
  )
ggsave(path_fig, plot = p, width = 14, height = 8, units = "cm")
```

\begin{center}
\includegraphics[width=0.8\linewidth  ]{`r path_fig`}
\end{center}


# Prediction Intervals: Derivation

- **Goal:** Obtain a prediction interval for a new response $Y_f$ at predictor values $X_f$.

- **Predicted value:**
  $$
  \hat{Y}_f = X_f' \hat{\beta}
  $$

- **Actual value:** $Y_f = X_f' \beta + \epsilon_f$, where $\epsilon_f \sim N(0, \sigma^2)$.



# Prediction Intervals: Derivation

- **Difference between actual and predicted values:**
  $$
  Z = Y_f - \hat{Y}_f
  $$

- **Expected value of $Z$:**
  $$
  E(Z) = E(Y_f - \hat{Y}_f) = X_f' \beta - X_f' E(\hat{\beta}) = 0
  $$

- **Variance of $Z$:**
  $$
  \text{Var}(Z) = \text{Var}(Y_f - \hat{Y}_f) = \sigma^2 + \sigma^2 X_f' (X'X)^{-1} X_f = \sigma^2 \left( 1 + X_f' (X'X)^{-1} X_f \right)
  $$



# Prediction Intervals: Derivation

- **Standardized variable:**
  $$
  \frac{Z}{\sqrt{\sigma^2 \left( 1 + X_f' (X'X)^{-1} X_f \right)}} \sim N(0,1)
  $$

- **Estimate $\sigma^2$ with $s^2$:**
  $$
  t = \frac{Y_f - \hat{Y}_f}{s \sqrt{1 + X_f' (X'X)^{-1} X_f}} \sim t_{n - k}
  $$

- **Prediction interval:**
  $$
  Y_f \in \hat{Y}_f \pm t_{\alpha/2, n - k} \cdot s \sqrt{1 + X_f' (X'X)^{-1} X_f}
  $$

# Variable Selection Motivation

- We may want fewer variables in our model for:
  - Interpretation purposes.
  - Practical application
  - Prediction accuracy:
    - Bias-variance trade-off.

- Aim: **Select the best subset of variables** that provides the optimal balance.

# The Combinatorial Challenge

- With $p$ predictors, there are $2^p$ possible models.

- **Example:** For $p = 100$, $2^{100} \approx 1.27 \times 10^{30}$ models.

- **Computationally infeasible** to evaluate all models.

# Greedy Algorithms for Variable Selection

- So, we adopt heuristic approaches to select variables.

- **Iterative, greedy optimization algorithms:**
  - *Iterative:* Start with a given solution and iteratively improve it.
  - *Greedy:* Always make the best immediate move without reconsideration.

- **Approaches:**
  - *Forward Selection*
  - *Backward Elimination*
  - *Stepwise Selection*



# Forward Selection

- Start with the intercept-only model.

- Iteratively add** the variable that provides the best improvement according to a criterion.

- Stop when no further improvement is possible.


# Model Selection Criteria

- **$R^2$:** Not suitable (always increases with more variables).

- **Adjusted $R^2$:** Adjusts for the number of predictors.

- **Akaike Information Criterion (AIC):**
  $$
  \text{AIC} = n \ln(\text{RSS}/n) + 2k
  $$

- **Bayesian Information Criterion (BIC):**
  $$
  \text{BIC} = n \ln(\text{RSS}/n) + \ln(n)k
  $$

- **Objective:** Minimize AIC or BIC.

# Example: Forward Selection in R

- **Simulate data with four variables: A, B, C, D.**

- **Variables A and B** are related to the response.

- **Variables C and D** are not related.

# R Code: Simulating Data

```{r}
set.seed(123)

# Sample size
n <- 100

# Predictor variables
A <- rnorm(n)
B <- rnorm(n)
C <- rnorm(n)
D <- rnorm(n)

# Response variable
Y <- 3 * A + 2 * B + rnorm(n)

# Data frame
data <- tibble(Y, A, B, C, D)
```

# R Code: Forward Selection

- **Load required package:**

```{r}
library(MASS) # For stepAIC function
```

- **Start with intercept-only model:**

```{r}
null_model <- lm(Y ~ 1, data = data)
full_model <- lm(Y ~ A + B + C + D, data = data)
```

- **Apply forward selection:**

```{r}
#| results: "hide"
forward_model <- stepAIC(null_model, scope = list(lower = null_model, upper = full_model), direction = "forward")
```

# Model fitting process

\begin{center}
\includegraphics[width=\linewidth]{_data_raw/img/forward_selection.png}
\end{center}

# Forward Selection: Results

```{r}
#| echo: false
coef(forward_model) |> signif(2)
```

- **Output shows that variables A and B are included.**

# Backward Elimination

- Start with the full model containing all predictors.

- Iteratively remove the variable that least contributes to the model.

- Stop when removing additional variables worsens the model.

# Stepwise Selection

- Combination of forward and backward methods.

- At each step:
  - Add the best variable.
  - Remove any variable that no longer improves the model.

- Allows variables to be reconsidered. (so, it's less greedy)

# R Code: Backward Elimination

- **Apply backward elimination:**

```{r}
#| results: "hide"
backward_model <- stepAIC(full_model, direction = "backward")
```

- **Selected variables:**

```{r}
coef(backward_model) |> signif(2)
```

- **Variables A and B remain in the model.**

# R Code: Stepwise Selection

- **Apply stepwise selection:**

```{r}
#| results: "hide"
stepwise_model <- stepAIC(
  null_model,
  scope = list(lower = null_model, upper = full_model),
  direction = "both"
)
```

- **Selected variables:**

```{r}
coef(stepwise_model) |> signif(2)
```

- **Again, variables A and B are selected.**

# Larger Example: Simulating High-Dimensional Data

- **Simulate data with 100 predictors, only 10 related to the response.**

```{r}
set.seed(123)
# Sample size
n <- 200; p <- 100

# Generate predictors
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("X", 1:p)

# True coefficients
beta <- c(rep(2, 5), rep(-2, 5), rep(0, p - 10))

# Response variable
Y <- X %*% beta + rnorm(n, sd = 0.05)

data_hd <- data.frame(Y, X) |> as_tibble()
```

# Apply forward, backward, and stepwise selection

```{r}
forward_model_hd <- stepAIC(
  lm(Y ~ 1, data = data_hd),
  scope = list(lower = ~1, upper = ~.),
  direction = "forward", trace = FALSE)
backward_model_hd <- stepAIC(
  lm(Y ~ ., data = data_hd),
  scope = list(lower = ~1, upper = ~.),
  direction = "backward", trace = FALSE
)
stepwise_model_hd <- stepAIC(
  lm(Y ~ 1, data = data_hd),
  scope = list(lower = ~1, upper = ~.),
  direction = "both", trace = FALSE
)
```

# Procedure performance

```{r}
#| echo: false
#| warning: false
#| message: false
fw_inc <- sum(names(coef(forward_model_hd)) %in% colnames(X))
bw_inc <- sum(names(coef(backward_model_hd)) %in% colnames(X))
sw_inc <- sum(names(coef(stepwise_model_hd)) %in% colnames(X))

fw_correct <- sum(names(coef(forward_model_hd)) %in% colnames(X)[1:10])
bw_correct <- sum(names(coef(backward_model_hd)) %in% colnames(X)[1:10])
sw_correct <- sum(names(coef(stepwise_model_hd)) %in% colnames(X)[1:10])

results_tbl <- tibble(
  Method = c("Full", "Forward", "Backward", "Stepwise"),
  `# Included` = c(100, fw_inc, bw_inc, sw_inc),
  `# Correct` = c(10, fw_correct, bw_correct, sw_correct),
  `R^2` = c(
    summary(lm(Y ~ ., data = data_hd))$r.squared,
    summary(forward_model_hd)$r.squared,
    summary(backward_model_hd)$r.squared,
    summary(stepwise_model_hd)$r.squared
  ),
  `R^2_adj` = c(
    summary(lm(Y ~ ., data = data_hd))$adj.r.squared,
    summary(forward_model_hd)$adj.r.squared,
    summary(backward_model_hd)$adj.r.squared,
    summary(stepwise_model_hd)$adj.r.squared
  ),
  AIC = c(
    AIC(lm(Y ~ ., data = data_hd)),
    AIC(forward_model_hd),
    AIC(backward_model_hd),
    AIC(stepwise_model_hd)
  ),
  BIC = c(
    BIC(lm(Y ~ ., data = data_hd)),
    BIC(forward_model_hd),
    BIC(backward_model_hd),
    BIC(stepwise_model_hd)
  )
)
results_tbl
```

# Estimating True Prediction Error

- **Simulate data where no predictors are related to the response.**

```{r}
set.seed(123)

n <- 150
p <- 100

# Generate predictors
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("X", 1:p)

# Response variable with no relation to predictors
Y <- rnorm(n)

# Data frame
data_null <- data.frame(Y, X)
```

# Applying Forward Selection

- **Use forward selection:**

```{r}
null_model <- lm(Y ~ 1, data = data_null)
full_model <- lm(Y ~ ., data = data_null)

forward_model <- stepAIC(
  null_model, scope = list(lower = null_model, upper = full_model), direction = "forward", trace = FALSE
)
```

We actually include `r sum(names(coef(forward_model)) != "(Intercept)")` variables in the model.
The mean squared error is then `r mean((data_null$Y - predict(forward_model, newdata = data_null))^2) |> signif(2)`.

# Observations on Forward Selection

- Our estimated error is much lower than it should be (it should be close to 1, the variance).
- Using the same data to evaluate the model as was used to select the model leads to under-estimating the error.

# Training-Test Split

- **Split data into training and test sets:**

```{r}
set.seed(123)
train_indices <- sample(1:n, size = floor(0.8 * n))

# Training data
train_data <- data_null[train_indices, ]

# Test data
test_data <- data_null[-train_indices, ]
```

# Evaluating Model Performance

- **Fit model on training data:**

```{r}
forward_model_cv <- stepAIC(
  lm(Y ~ 1, data = train_data),
  scope = list(lower = ~1, upper = ~.),
  direction = "forward", trace = FALSE
)
```

- **Predict on test data:**

```{r}
predictions <- predict(forward_model_cv, newdata = test_data)
```

- **Calculate Mean Squared Error (MSE):**

```{r}
mse_test <- mean((test_data$Y - predictions)^2)
```

The error is now `r mse_test |> signif(2)`, whereas when calculated on in-sample data it was `r mean((data_null$Y - predict(forward_model, newdata = data_null))^2) |> signif(2)`.


# Observations on error rate estimation

- Test MSE was closer to the true error than the in-sample error.

# Summary

- **Prediction intervals** account for uncertainty in both parameter estimation and future observations.

- **Variable selection** helps in building parsimonious models.

- **Selection criteria** like AIC and BIC balance model fit and complexity.

- **Forward, backward, and stepwise selection** are practical methods for variable selection.

- **Training-test split error estimation** helps in estimating true prediction error.