---
title: Principal component analysis
author: Miguel Rodo
bibliography: zotero.bib
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: preamble.tex
nocite: |
    @johnson_wichern07
---

```{r }
#| include: false
library(tibble)
library(ggplot2)
library(DataTidy23RodoHonsMult)
for (x in list.files("R", full.names = TRUE)) {
  source(x)
}
```

# Context

- Multivariate analysis helps us uncover patterns and relationships within datasets containing multiple variables.
- **Example**: Meteorologists analyze temperature, humidity, wind speed, and more to understand weather systems.
  - Possible questions:
    - What sets of variable(s) are correlated?
    - Can we predict the likelihood of rain based on these variables?
    - What variables most distinguish between different weather patterns?

# Basis of approach

- **Linear combinations**:

$$
\symbf{x}_1\symbf{u}_1 + \symbf{x}_2\symbf{u}_2 + \ldots + \symbf{x}_p\symbf{u}_p
$$

- Why linear combinations?
  - Simplicity >> strong theoretical results >> robustness, speed and interpretability

# Types of multivariate analysis

- $\symbf{X} \sim \symbf{X}$: PCA, factor analysis, correspondence analysis (count data)
  - **Examples**: identify correlates sets of variables
- $\symbf{Y} \sim \symbf{X}$: Multiple/multivariate regression, canonical correlation analysis, discrimination and classification

# Principal components analysis

- Principal components analysis (PCA) identifies the directions of greatest variation, allowing you to:
  - Represent data more compactly, and
  - Interpret relationships between variables.

- PCA is a linear technique with an algebraic solution, so it is fast to fit.

# Example applications

- **Market analysis**: Identify correlated stocks to ensure a diversified portfolio
- **Image compression**: Represent images with fewer pixels while preserving accuracy
- **Bioinformatics**: Identify sets of co-expressed genes responsible for different functions in the body

# Two-variable example

:::: {.columns}

::: {.column width="50%"}

- Suppose we have a dataset with two correlated variables, perhaps weight and height.
- Then, the scatterplot of these may look something like this:

```{r}
set.seed(2106)
plot_tbl <- tibble::tibble(
  height = rnorm(100, 160, 10)
) |>
  dplyr::mutate(
    weight = height / 2.5 + rnorm(100, 0, 3.5 * height / 160)
  ) |>
  dplyr::mutate(
    across(everything(), function(x) (x - mean(x)) / sd(x))
  ) |>
  dplyr::arrange(height)
hw_tbl <- plot_tbl
p <- ggplot(
  aes(x = height, y = weight),
  data = plot_tbl
) +
  geom_vline(
    xintercept = 0, col = "gray50", linetype = "solid"
  ) +
  geom_hline(
    yintercept = 0, col = "gray50", linetype = "solid"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme( 
    panel.background = element_rect(fill = "white"), 
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Height",
    y = "Weight"
  ) +
  geom_point(alpha = 0.85, col = "dodgerblue")
path_p_height_weight <- projr::projr_path_get("cache", "fig", "p-height-weight.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_height_weight,
  units = "cm",
  base_height = 5,
  base_width = 5
)
```

\centering

```{r}
knitr::include_graphics(path_p_height_weight)
```

:::

::: {.column width="50%"}

- We could combine these two variables into a single variable, representing size: $a_1 \times \text{height} + a_2 \times \text{weight}$.

:::

::::

# Two-variable example (cont.) {.smaller}

:::: {.columns}

::: {.column width="50%"}

- The new variable is then the *length and sign of the projection*.
- For example, suppose we have the point $[1.2, 0.3]$:

```{r}
set.seed(2106)
plot_tbl <- tibble::tibble(
  height = rnorm(100, 160, 10)
) |>
  dplyr::mutate(
    weight = height / 2.5 + rnorm(100, 0, 3.5 * height / 160)
  ) |>
  dplyr::mutate(
    across(everything(), function(x) (x - mean(x)) / sd(x))
  ) |>
  dplyr::arrange(height)
proj_list <- list(c(1 / sqrt(2), 1 / sqrt(2)), c(1, 0), c(0, 1))[1]
plot_tbl_ind <- tibble::tibble(
  height = 1.2,
  weight = 0.3
)
plot_tbl_proj_ind <- proj_list |>
  purrr::map_df(~ mutate_proj(plot_tbl_ind, .x))

p <- ggplot(
  data = plot_tbl
) +
  geom_vline(
    xintercept = 0, col = "gray50", linetype = "solid"
  ) +
  geom_hline(
    yintercept = 0, col = "gray50", linetype = "solid"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Height",
    y = "Weight"
  ) +
  geom_point(aes(x = height, y = weight), alpha = 0.1, col = "dodgerblue") +
  geom_point(
    data = plot_tbl_proj_ind,
    aes(x = height, y = weight), col = "mediumseagreen"
  ) +
  geom_segment(
    data = plot_tbl_proj_ind,
    aes(x = height, y = weight, xend = proj_x, yend = proj_y, group = proj),
    col = "gray25", arrow = arrow(length = unit(0.3, "cm")), linetype = "solid"
  ) +
  geom_abline(
    data = plot_tbl_proj_ind |>
      dplyr::mutate(slope = a2 / a1) |>
      dplyr::filter(slope %in% c(0,1)),
    aes(intercept = 0, slope = slope),
    col = "purple"
  ) +
  coord_cartesian(xlim = c(0, 2.5), ylim = c(0, 2.5)) +
  ggrepel::geom_text_repel(
    data = plot_tbl_proj_ind |>
      dplyr::mutate(a1 = a1 |> signif(2), a2 = a2 |> signif(2)),
    aes(x = a1, y = a2, label = proj),
    nudge_x = 0.1, nudge_y = 0.1
  )
path_p_height_weight_ind <- projr::projr_path_get("cache", "fig", "p-height-weight-ind.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_height_weight_ind,
  units = "cm",
  base_height = 5,
  base_width = 5
)
```

\centering

```{r}
#| results: asis
knitr::include_graphics(path_p_height_weight_ind)
```

:::

::: {.column width="50%"}

- If we set $\symbf{a} = [1/\sqrt{2}, 1/\sqrt{2}]$, then the new variable is $1/\sqrt{2} \times 1.2 + 1/\sqrt{2} \times 0.3$.
- We go from the matrix of vairables

```{r}
tibble::tibble(height = 1.2, weight = 0.3) |> as.data.frame()
```

- to

```{r}
tibble::tibble(size = 1.2 / sqrt(2) + 0.3 / sqrt(2)) |> as.data.frame()
```

:::

::::

# How do we choose $a_1$ and $a_2$? {.smaller}

- Consider projecting onto the lines $[0.5, 2]$, $[1, -1]$ and $[1, 1]$:

```{r}
proj_list <- list(c(0.5, 2), c(1, 1), c(1, -1))
plot_tbl_proj <- purrr::map_df(proj_list, ~mutate_proj(plot_tbl, .x))

p <- ggplot(
  data = plot_tbl_proj
) +
  geom_vline(
    xintercept = 0, col = "gray50", linetype = "solid"
  ) +
  geom_hline(
    yintercept = 0, col = "gray50", linetype = "solid"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme( 
    panel.background = element_rect(fill = "white"), 
    plot.background = element_rect(fill = "white")
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Height",
    y = "Weight"
  ) +
  geom_point(aes(x = height, y = weight), alpha = 0.85, col = "dodgerblue") +
  geom_segment(
    data = plot_tbl_proj |>
      dplyr::group_by(proj) |>
      dplyr::slice(seq(1, 100, length.out = 5) |> round()) |>
      dplyr::ungroup(),
    aes(x = height, y = weight, xend = proj_x, yend = proj_y, group = proj),
    col = "gray25", arrow = arrow(length = unit(0.3, "cm")), linetype = "solid"
  ) +
  geom_abline(
    data = plot_tbl_proj |>
      dplyr::group_by(proj) |>
      dplyr::slice(1) |>
      dplyr::ungroup() |>
      dplyr::mutate(slope = a2 / a1),
    aes(intercept = 0, slope = slope),
  ) +
  facet_wrap(~proj, ncol = 3)
path_p_proj <- projr::projr_path_get("cache", "fig", "p-proj.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_proj,
  units = "cm",
  base_height = 6,
  base_width = 12
)
```

\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{`r path_p_proj`}
\end{figure}

::::{.columns}

::: {.column width="50%"}
- Height and weight increase together, so $a_1$ and $a_2$ should both be positive.
- Beyond that, we want to choose $a_1$ and $a_2$ so that the projection captures the most variation in the data.
:::

::: {.column width="50%"}
- The variability of the (lengths of the) projections is 1.3 (0.5, 2), 0.3 (1, -1) and 1.7 (1, 1), respectively.
- The projection onto $[1, 1]$ captures the most variation.
:::

::::

# Further motivation for capturing variation

- On its own grounds, imagine that you could retain only one of two variables: one has zero variance (all the observations are the same) and the other has some positive variance. Clearly, you would choose the latter.

\pause

- Later, we will see that capturing maximal variation:
  - Minimises the distance from the original data to the projection.
  - Allows us to reconstruct the original variables with minimal error.

# Objective function: the first principal component

- Assume that our data $\symbf{X} \in \mathcal{R}^p$ are distributed with mean $\symbf{\symbf{\mu}}$ and covariance matrix $\symbf{\Sigma}$.
- We seek to find the linear combination of the variables that captures the most variation.
- For the first principal component, we seek to find $\symbf{a}_1$ such that

$$
\text{Var}(\symbf{a}_1'\symbf{X}) = \symbf{a}_1^T\symbf{\Sigma}\symbf{a}_1
$$

- is maximized, subject to the constraint that $||\symbf{a}_1|| = 1$.
  - This is necessary as otherwise we could meaninglessly increase the variation without bound by making $\symbf{a}_1$ longer.
- Applied to the previous example, the first principal component would be a linear combination of height and weight that captures the most variation.

# Objective function: the second principal component

- The second principal component is the linear combination of the variables that captures the most variation, subject to the constraint that it is orthogonal to the first principal component.
- In other words, for the second principal component we seek to find $\symbf{a}_2$ such that

$$
\text{Var}(\symbf{a}_2'\symbf{X}) = \symbf{a}_2^T\symbf{\Sigma}\symbf{a}_2
$$

- is maximized, subject to the constraints that $||\symbf{a}_2|| = 1$ and $\symbf{a}_2^T\symbf{a}_1 = 0$.


# Objective function for all principal components

- For the $k$-th principal component (for $k\leq p$), we seek to find $\symbf{a}_k$ such that

$$
\text{Var}(\symbf{a}_k'\symbf{X}) = \symbf{a}_k^T\symbf{\Sigma}\symbf{a}_k
$$

- is maximized, subject to the constraints that $||\symbf{a}_k|| = 1$ and $\symbf{a}_k^T\symbf{a}_j = 0$ for $j < k$.

\pause

- Up until this point, we've explored various options for $\symbf{a}_k$ without any method. We'll now discuss how to compute these principal components.
- First, we'll discuss a more general result.

# Weights for variance-maximising linear combinations {.smaller}

Let $\symbf{X}:\symbf{p}\times 1$ be the random vector with variance-covariance matrix $\symbf{\Sigma}:p\times p$, which has the $i$-th largest eigenvalue as $\lambda_i$ and associated eigenvector $\symbf{e}_i$.

Then the $i$-th principal component (defined as before) is given by 

$$
Y_i = \symbf{e}_i'\symbf{X},
$$

implying that

\begin{align*}
\mathrm{Var}[Y_i] &= \lambda_i \;\forall \; i \in \{1,2, \ldots, p\}, \; \mathrm{and} \\
\mathrm{Cov}[Y_i, Y_j] &= 0 \; \mathrm{for} \; i\neq j.
\end{align*}

# A note on terminology

- Principal components (the $\symbf{Y}_i$'s) may also be referred to as "scores".
- The coefficient vectors $\symbf{a}_i$ may also be referred to as "loadings" (thinking algebraically) or as "principal axes"/"principal directions" (thinking geometrically).

# Example {.smaller}

::::{.columns}

:::{.column width="50%"}

Suppose that we have the following variance-covariance matrix:

```{r}
#| echo: true
vcov_mat <- matrix(
  c(1, 0.9, 0.9, 0, 0, 0, 0,
    0.9, 1, 0.9, 0, 0, 0, 0,
    0.9, 0.9, 1, 0, 0, 0, 0,
    0, 0, 0, 1, 0.9, 0.9, 0.9,
    0, 0, 0, 0.9, 1, 0.9, 0.9,
    0, 0, 0, 0.9, 0.9, 1, 0.9,
    0, 0, 0, 0.9, 0.9, 0.9, 1),
  byrow = TRUE,
  nrow = 7
)
colnames(vcov_mat) <- paste0("V", 1:7)
rownames(vcov_mat) <- paste0("V", 1:7)
```

:::

:::{.column width="50%"}

```{r}
vcov_mat
```

:::

::::

# Example (cont.) {.smaller}

::::{.columns}

:::{.column width="50%"}

- We apply the eigen decomposition to the variance-covariance matrix:

```{r}
#| echo: true
eig_obj <- eigen(vcov_mat)
```

- to obtain the following eigenvectors:

```{r}
#| echo: true
eig_vec_mat <- eig_obj$vectors
rownames(eig_vec_mat) <- paste0("V", 1:7)
colnames(eig_vec_mat) <- paste0("PC", 1:7)
```

:::

:::{.column width="50%"}

```{r}
eig_vec_mat |> signif(2)
```

:::

::::

# Example (cont.) {.smaller}

::::{.columns}

:::{.column width="50%"}

- By the variance formula, these linear combinations have the following variances:

```{r}
#| echo: true
(eig_vec_mat |> t()) %*%
  vcov_mat %*%
  (eig_vec_mat) |>
  diag() |>
  signif(2)
```

:::

:::{.column width="50%"}

- These are the same as the eigenvalues:

```{r}
#| echo: true
eig_obj$values |>
  signif(2) |>
  setNames(paste0("PC", 1:7))
```

- Clearly, due to the correlation structure, the orthogonality constraint implies that only two PCs have meaningful variance.

:::

::::

# Total variance {.smaller}

The sum of variances of the original variables is equal to the sum of the eigenvalues of $\symbf{\Sigma}$, which is also equal to the sum of variances of the principal components:

$$
\sum_{i=1}^{p} \sigma_{ii} = \sum_{i=1}^{p} \text{Var}(X_i) = \sum_{i=1}^{p} \lambda_i = \sum_{i=1}^{p} \text{Var}(Y_i).
$$

# Contribution per component

We've seen before that

$$
\text{trace}(\symbf{\Sigma}) = \sum \lambda_i = \sum \text{Var}(Y_i).
$$

This implies that the proportion of the total variance due to the $k^{th}$ principal component is

$$
\frac{\lambda_k}{\sum \lambda_i },\quad k=1,...p,
$$

and (due to being uncorrelated) the proportion of variance accounted for by the first $r$ principal components is given by

$$
\frac{\sum_{i=1}^r \lambda_i }{\sum \lambda_i }.
$$

# The scree plot

:::: {.columns}

::: {.column width="50%"}

- The scree plot is a graphical representation of the proportion of variance accounted for by each principal component.
- It plots $(i, \lambda_i/\sum_{j=1}^{p} \lambda_j)$ for $i = 1, \ldots, p$.
- Typically, we look for the "elbow" in the plot, which indicates the number of principal components to retain.
- Beyond this point, each additional principal component captures little additional variation.

:::

::: {.column width="50%"}

```{r}
plot_tbl <- tibble::tibble(
  pc = 1:7,
  prop_var = eig_obj$values / sum(eig_obj$values)
)
plot_tbl_point <- purrr::map_df(
  seq(5.5, 5.502, length.out = 9), function(x) {
    plot_tbl |>
      dplyr::filter(pc == 2) |>
      dplyr::mutate(size_point = x)
  }
)
p <-  ggplot(
  data = plot_tbl
  ) +
    geom_line(
      aes(x = pc, y = prop_var),
      stat = "identity",
      colour = "dodgerblue"
    ) +
    geom_point(
      data = plot_tbl_point |> dplyr::slice(5:9),
      aes(x = pc, y = prop_var, size = size_point),
      colour = "mediumseagreen",
      alpha = 0.9,
      shape = 1,
      show.legend = FALSE
    ) +
    geom_point(
      aes(x = pc, y = prop_var),
      colour = "dodgerblue",
      alpha = 0.9,
      size = 1.5
    ) +
    labs(
      x = "Principal component index",
      y = "Proportion of variance"
    ) +
    cowplot::theme_cowplot() +
    cowplot::background_grid(major = "xy") +
    theme(
      panel.background = element_rect(fill = "white"),
      plot.background = element_rect(fill = "white")
    ) +
    scale_x_continuous(
      breaks = 1:7
    )
path_p_scree <- projr::projr_path_get("cache", "fig", "p-scree.pdf")
cowplot::save_plot(
  plot = p,
  filename = path_p_scree,
  units = "cm",
  base_height = 7,
  base_width = 8 * (7/6)
)
```

```{r}
#| results: asis
knitr::include_graphics(path_p_scree)
```

:::

::::

# Variance standardisation

- Before applying PCA, one can standardise the variances to have unit variance.
- This may be appropriate when:
  - the variances have different units (e.g. weight in kg vs annual income in rands), or
  - the variances have different scales (e.g. some genes are extremely rare whereas others are abundantly expressed), or
  - we wish to eliminate any effect of differences in variance on downstream analyses, e.g. when applying penalised regression.
- Similar comments go for logging the data, or applying other transformations.
- All results go through in exactly the same way, with adjustments made in interpreting exactly what variables the principal components capture.

# Standardisation example

::::{.columns}

:::{.column width="50%"}

We have the following covariance matrix:

```{r}
cov_mat <- matrix(
  c(1, 4, 4, 100), nrow = 2
)
cov_mat
```

which produces the following eigendecomposition:

```{r}
eigen(cov_mat) |> lapply(function(x) signif(x, 2))
```

:::

:::{.column width="50%"}

The associated correlation matrix:

```{r}
diag_mat_sd_inv <- diag(diag(cov_mat), nrow = nrow(cov_mat)) |> sqrt() |> solve()
rho_mat <- diag_mat_sd_inv %*% cov_mat %*% diag_mat_sd_inv
rho_mat 
```

has the following eigendecomposition:

```{r}
eigen(rho_mat) |> lapply(function(x) signif(x, 2))
```

The eigenvectors are in very different directions.

:::

::::

# Example

:::: {.columns}

::: {.column width="57%"}

We read in data of measurements Painted Turtles [@jolicoeur_mosimann60]:

```{r}
#| eval: false
#| echo: true
if (!requireNamespace(
  "remotes", quietly = TRUE
  )) {
  install.packages("remotes")
}
remotes::install_github(
  "MiguelRodo/DataTidy23RodoHonsMult@2024"
)  
```

:::

::: {.column width="43%"}

```{r}
#| echo: true
data(
  "data_tidy_turtle",
  package = "DataTidy23RodoHonsMult"
)
data_tidy_turtle
```

:::

::::

# Measurements highly correlated

::::{.columns}

:::{.column width="50%"}

```{r}
#| include: false
plot_tbl_turtle <- data_tidy_turtle |>
  tidyr::pivot_longer(
    length:width,
    names_to = "x_var",
    values_to = "x_val"
  )
p <- ggplot(
  plot_tbl_turtle,
  aes(x = x_val, y = height, col = gender)
) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(
    values = c(male = "dodgerblue", female = "mediumseagreen")
   ) +
  facet_wrap(~x_var, scales = "free", ncol =1 ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme( 
    panel.background = element_rect(fill = "white"), 
    plot.background = element_rect(fill = "white")
  ) +
  labs(
    x = "Measurement",
    y = "Height"
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  theme(
    strip.background = element_rect(fill = "white", colour = "gray25")
  )
path_p_turtle <- projr::projr_path_get("cache", "fig", "p-turtle.png")
cowplot::save_plot(
  plot = p,
  filename = path_p_turtle,
  units = "cm",
  base_height = 7,
  base_width = 5
)
```

```{r}
#| results: "asis"
knitr::include_graphics(path_p_turtle)
```

:::

:::{.column width="50%"}

- Due to the correlation, we may likely summarise the data very well using principal components.
- We'll use the `prcomp` function in R, and do it from first principles.

:::

::::

# Using the `prcomp` function I {.smaller}

::::{.columns}

:::{.column width="70%"}

```{r}
#| echo: true
pr_obj <- prcomp(
  ~ log(length) + log(width) + log(height),
  data = data_tidy_turtle,
  retx = TRUE
)
pr_obj
```

:::

:::{.column width="30%"}

- The first principal component has a markedly higher standard deviation.
- It is weighted roughly evenly across the variables, indicating it is a general "size" variable.

:::

::::

# Using the `prcomp` function II {.smaller}

::::{.columns}

:::{.column width="70%"}

```{r}
#| echo: true
summary(pr_obj)
```

:::

:::{.column width="30%"}

- The `summary` function provides the per-component standard deviation and variance, as well as the cumulative variance.

:::

::::

# Using the `prcomp` function III {.smaller}

::::{.columns}

:::{.column width="70%"}

```{r}
#| echo: true
plot(pr_obj)
```

:::

:::{.column width="30%"}

- The `plot` command provides the scree plot.
- Clearly, unless we're specifically interested in non-size variables we should only retain the first principal component.

:::

::::

# Using the `prcomp` function IV {.smaller}

Here are the principal components:

```{r}
#| echo: true
pr_obj$x |> head() |> signif(2)
```

# Interpretation guideliness {.smaller}

```{r}
#| echo: false
example_mat <- matrix(
  c(
    0.8, 0.6, 0.01, 001,
    0.01, 0.3, 0.3, -0.9055
    ),
  nrow = 2,
  byrow = TRUE
)
rownames(example_mat) <- c("PC1", "PC2")
colnames(example_mat) <- c("CO2", "NO", "O3", "Wind speed")
example_mat
```

- **Summarize, Don't Repeat:** Avoid simply listing coefficients from tables. Instead, concisely summarize the key insights in a few sentences. 
- **Ignore Trivial Coefficients:** Very small coefficients (e.g., 0.001) are usually negligible and can be omitted from your interpretation.
- **Use Coefficients Judiciously:** Mention coefficients in parentheses when they strengthen your interpretation.
- For example:
  - "Component one is driven by O3 (0.08) and NO (0.54))."
  - "Component two contrasts O3  and NO (both 0.3) with Wind speed (-0.9)."

# References

