---
title: "Regression Week IV: Model Checking"
subtitle: "Model Checking"
author: "Yovna Junglee"
institute: "Department of Statistical Sciences - University of Cape Town"
bibliography: zotero.bib
format: 
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: preamble.tex
execute:
  echo: true
---

```{r}
library(tibble)
library(ggplot2)
```

# So far...

1. **Formulated** a linear model.
2. **Parameter estimation** using least squares.
3. Constructed **confidence intervals** for the parameters.
4. Performed **hypothesis testing** and **ANOVA**.

**Next:** **Model checking** to validate our model assumptions.

# Proportion of variation explained by the model

- The proportion of variation explained by the model is given by the **coefficient of determination**, $R^2$:

  $$
  R^2 = \frac{SSR}{SST} = 1 - \frac{SSE}{SST},
  $$

  where

  - $SSR$: Regression sum of squares
    - $\sum_{i=1}^n (\hat{Y}_i - \bar{Y})^2$
  - $SSE$: Error sum of squares
    - $\sum_{i=1}^n (Y_i - \hat{Y}_i)^2$
  - $SST$: Total sum of squares
    - $\sum_{i=1}^n (Y_i - \bar{Y})^2$

- It is a sensible formula as $SST = SSR + SSE$.
- It is also known as the coefficient of determination.


# Coefficient of Determination - Example

To understand the **coefficient of determination ($R^2$)**, let's consider an example where we fit models with an increasing number of predictors.

```{r}
# Simulate data
set.seed(123)
n <- 100
X1 <- rnorm(n)
X2 <- rnorm(n)
X3 <- rnorm(n)
epsilon <- rnorm(n, sd = 2)
Y <- 3 + 2 * X1 + epsilon

# Fit models with increasing predictors
model1 <- lm(Y ~ 1)           # Intercept only
model2 <- lm(Y ~ X1)          # One predictor
model3 <- lm(Y ~ X1 + X2)     # Two predictors
model4 <- lm(Y ~ X1 + X2 + X3) # Three predictors
mod_list <- list(model1, model2, model3, model4)
```

# Coefficient of Determination - Example

- The $R^2$ value tends to increase as we add more predictors:

```{r}
# Compute R-squared values
R2_values <- lapply(mod_list, function(x) summary(x)$r.squared)
R2_values |> sapply(function(x) signif(x, 4))
```

```{r}
# Compute predicted values
Y_hat <- lapply(mod_list, function(x) predict(x))
# Compute overall mean
```

# Coefficient of Determination - Example

```{r}
plot(1:4, R2_values, type = "b", pch = 19,
     xlab = "Model", ylab = expression(R^2),
     main = expression("R"^2 * " Values for Models with Increasing Predictors"),
     xaxt = "n", ylim = c(0, 1))
axis(1, at = 1:4, labels = c("Intercept Only", "X1", "X1 + X2", "X1 + X2 + X3"))
```

**Observation:** $R^2$ increases as we add more variables, indicating an improved fit.

# Coefficient of Determination

1. The example demonstrates that adding variables improves the model fit.
2. Measuring the fit via the correlation between fitted and true $Y$ values is sensible.
3. The correlation coefficient $R$ is:

   $$
   R = \sqrt{\frac{\sum_{j=1}^n \hat{Y}_j^2 - n\bar{Y}^2}{\sum_{j=1}^n Y_j^2 - n\bar{Y}^2}} = \sqrt{\frac{\hat{\beta}'X'Y - n\bar{Y}^2}{Y'Y - n\bar{Y}^2}}
   $$
   
4. $R$ ranges between $-1$ and $1$.

# Coefficient of Determination Continued

1. We often use the square of $R$, the **coefficient of determination ($R^2$)**:

   $$
   R^2 = \frac{SSR}{SST} = 1 - \frac{SSE}{SST}
   $$
   
   - $SSR$: Regression sum of squares.
   - $SSE$: Error sum of squares.
   - $SST$: Total sum of squares.

2. $R^2$ ranges from 0 to 1.
3. An $R^2$ close to 1 indicates a good fit.
4. An $R^2$ close to 0 suggests a poor fit.
5. A higher $R^2$ does not guarantee significant $\beta$ estimates.
6. $R^2$ is a valid measure primarily for linear models.

**Interpretation:** $R^2$ represents the proportion of variance in $Y$ explained by the model.

# Some More Comments

1. **Adding variables increases $R^2$**, even if they are not significant—this can be misleading.
2. Use the **Adjusted $R^2$** ($R^2_{\text{adj}}$) to account for the number of predictors:

   $$
   R^2_{\text{adj}} = 1 - \frac{SSE/(n - k)}{SST/(n - 1)} = 1 - (1 - R^2)\left(\frac{n - 1}{n - k}\right)
   $$
   
   - $n$: Number of observations.
   - $k$: Number of parameters (including the intercept).

3. $R^2_{\text{adj}} \leq R^2$.
4. When comparing models, choose the one with the highest $R^2_{\text{adj}}$.

# Example

Calculate $R^2$ and $R^2_{\text{adj}}$ for our models:

```{r}
# Compute Adjusted R-squared values
adj_R2_values <- c(
  summary(model1)$adj.r.squared,
  summary(model2)$adj.r.squared,
  summary(model3)$adj.r.squared,
  summary(model4)$adj.r.squared
)

# Combine results
R2_results <- data.frame(
  Model = c("Intercept Only", "X1", "X1 + X2", "X1 + X2 + X3"),
  R_squared = round(R2_values, 4),
  Adjusted_R_squared = round(adj_R2_values, 4)
)

print(R2_results)
```

**Observation:** Adjusted $R^2$ may decrease if added predictors do not improve the model significantly.

# Model Checking

## Model Assumptions

For our linear regression model, we assume:

1. **Linearity**: $E(e_j) = 0$ for all $j$.
2. **Homoscedasticity**: $E(e_j^2) = \sigma^2$ for all $j$.
3. **Independence**: $E(e_j e_k) = 0$ for $j \neq k$.
4. **Normality**: $e_j \sim N(0, \sigma^2)$ for all $j$.

**Goal:** Validate these assumptions through residual analysis.

# Model Checking Plots (Using the `mtcars` Dataset)

We will use the `mtcars` dataset to perform model checking.

```{r}
# Load dataset
data(mtcars)

# Fit multiple linear regression model
fit <- lm(mpg ~ wt + hp + disp + drat, data = mtcars)
```

### Residuals vs Fitted Plot

```{r}
# Residuals vs Fitted
plot(fit$fitted.values, fit$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red")
```

- **Analysis**: Check for patterns. A funnel shape suggests heteroscedasticity.
- **Observation**: Residuals appear randomly scattered around zero.

# Analysis of Residuals

## Outliers vs Influential Observations

- **Outlier**: An observation not well fitted by the model (large residual).
- **Influential Observation**: Significantly affects model estimates (high leverage).

We use statistical measures to detect them.

# Estimated Residuals

Residuals are calculated as:

$$
\hat{e} = Y - \hat{Y} = (I - H)Y = MY
$$

- $M = I - H$ is idempotent and symmetric.
- $H = X(X'X)^{-1}X'$ is the hat matrix.

# Analysis of Residuals - Outliers

Variance of residuals:

$$
\text{Var}(\hat{e}_i) = \sigma^2(1 - h_i)
$$

Standardized residuals:

$$
r_i = \frac{\hat{e}_i}{s\sqrt{1 - h_i}}
$$

- $s^2 = \frac{\hat{e}'\hat{e}}{n - k}$
- $h_i$: Leverage of observation $i$

**Rule of Thumb:** $|r_i| > 2$ indicates potential outliers.

```{r}
# Standardized residuals
std_residuals <- rstandard(fit)

# Plot
plot(fit$fitted.values, std_residuals,
     xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Standardized Residuals vs Fitted Values")
abline(h = c(-2, 0, 2), col = "red", lty = 2)
```

# Analysis of Residuals - Externally Studentized Residuals

Externally studentized residuals:

$$
t_i = \frac{\hat{e}_i}{s_{(i)}\sqrt{1 - h_i}}
$$

- $s_{(i)}$: Standard error without observation $i$

**Use:** Detect outliers more accurately.

```{r}
# Externally studentized residuals
stud_residuals <- rstudent(fit)

# Plot
plot(fit$fitted.values, stud_residuals,
     xlab = "Fitted Values", ylab = "Studentized Residuals",
     main = "Externally Studentized Residuals vs Fitted Values")
abline(h = c(-2, 0, 2), col = "red", lty = 2)
```

**Observation:** Points beyond $\pm2$ are potential outliers.

# Analysis of Residuals - Influential Points

Leverage ($h_i$):

$$
h_i = \text{Diagonal of } H
$$

- Average $h_i$: $\frac{k}{n}$
- **Rule of Thumb:** $h_i > \frac{2k}{n}$ indicates high leverage.

```{r}
# Leverage values
leverages <- hatvalues(fit)

# Plot
plot(leverages, type = "h",
     xlab = "Observation", ylab = "Leverage",
     main = "Leverage Values")
abline(h = 2 * length(coef(fit)) / nrow(mtcars), col = "red", lty = 2)
```

**Observation:** Identify observations with high leverage.

# Cook's Distance and Modified Cook's Statistic

**Cook's Distance ($D_i$)**:

$$
D_i = \left( \frac{t_i^2}{k} \right) \frac{h_i}{1 - h_i}
$$

- Measures influence of observation $i$.
- **Cut-off:** $D_i > \frac{4}{n - k - 1}$

```{r}
# Cook's Distance
cooks_d <- cooks.distance(fit)

# Plot
plot(cooks_d, type = "h",
     xlab = "Observation", ylab = "Cook's Distance",
     main = "Cook's Distance Plot")
abline(h = 4 / (nrow(mtcars) - length(coef(fit))), col = "red", lty = 2)
```

**Observation:** Points above the red line are influential.

# Modified Cook's Distance (Atkinson's Method)

**Atkinson's Statistic ($A_i$)**:

$$
A_i = \sqrt{(n - k)D_i} = \sqrt{\left( \frac{n - k}{k} \right) \frac{h_i}{1 - h_i}} |t_i|
$$

- **Cut-off values:** 2 or 3.

```{r}
# Atkinson's Statistic
A_i <- sqrt((nrow(mtcars) - length(coef(fit))) * cooks_d)

# Plot
plot(A_i, type = "h",
     xlab = "Observation", ylab = "Atkinson's Statistic",
     main = "Atkinson's Plot")
abline(h = c(2, 3), col = c("blue", "red"), lty = 2)
legend("topright", legend = c("Cut-off at 2", "Cut-off at 3"),
       col = c("blue", "red"), lty = 2)
```

**Observation:** Points above the lines are potentially influential.

# Summary

- **Outliers**: Large residuals; do not fit well.
- **Influential Points**: Significantly impact model parameters.
- **Leverage**: Indicates potential influence based on predictor values.
- **Studentized Residuals**: Detect outliers.
- **Cook's Distance**: Identifies influential observations.
- **Atkinson's Statistic**: Alternative measure for influence.
