---
title: "Bootstrapping"
author: "Miguel Rodo"
institute: "Department of Statistical Sciences - University of Cape Town"
bibliography: zotero.bib
nocite: |
  @hesterberg14
format: 
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: preamble.tex
execute:
  echo: true
---


```{r}
#| include: false
library(tibble)
library(ggplot2)
theme_set(
  cowplot::theme_cowplot() + 
    theme(
      plot.background = element_rect(fill = "white"),
      panel.background = element_rect(fill = "white"),
    )
  )
```

# Introduction

Suppose that we have skewed data

```{r}
x_vec <- seq(0.0001, 10, length.out = 1e3)
skew_tbl <- tibble(
  x = x_vec,
  y = dexp(x_vec, rate = 0.1)^1.5
)
```

```{r}
#| include: false
p <- ggplot(skew_tbl) +
  geom_line(aes(x = x, y = y), color = "dodgerblue", size = 1) +
  labs(x = "Value", y = "Frequency") +
  cowplot::background_grid(major = "xy", minor = "none")
path_fig <- projr::projr_path_get(
  "cache", "fig", "w6", "p-init_example.pdf"
)
ggsave(
  path_fig, plot = p, width = 10, height = 7, dpi = 150,
  units = "cm"
)
```

\centering
\begin{centering}
\includegraphics[width=0.4\textwidth]{`r path_fig`}
\end{centering}

# The distribution of the mean will tend to be skewed as well, especially at lower sample sizes

```{r}
set.seed(4)
mean_tbl <- purrr::map_df(
  c(5, 10, 20, 40, 100, 1e3), function(x) {
    set.seed(3)
    mean_vec <- replicate(1e3, {
      x_vec <- rexp(x, rate = 0.1)^1.5
      mean(x_vec)
    })
    tibble(
      sample_size = x,
      mean = mean_vec
    )
  }
)
```

# The distribution of the mean will tend to be skewed as well, especially at lower sample sizes

```{r}
#| include: false
mean_tbl_norm <- purrr::map_df(
  c(5, 10, 20, 40, 100, 1e3), function(x) {
    set.seed(3)
    mean_vec <- mean_tbl |>
      dplyr::filter(sample_size == x) |>
      dplyr::pull(mean)
    mean_norm <- mean(mean_vec)
    sd_norm <- sd(mean_vec)
    x_vec <- seq(min(mean_vec), max(mean_vec), length.out = 1e3)
    y_vec <- dnorm(x_vec, mean = mean_norm, sd = sd_norm)
    tibble(
      sample_size = x,
      x = x_vec,
      y = y_vec
    )
  }
)
p <- ggplot(
  mean_tbl, aes(x = mean, fill = factor(sample_size))
) +
  geom_line(
    data = mean_tbl_norm,
    aes(x = x, y = y, group = sample_size),
    color = "black", size = 1,
    inherit.aes = FALSE
  ) +
  geom_density(alpha = 0.7) +
  labs(x = "Mean", y = "Density") +
  scale_fill_viridis_d(option = "plasma", name = "Sample Size") +
  facet_wrap(~ sample_size, scales = "free") +
  cowplot::background_grid(major = "xy", minor = "none") +
  theme(legend.position = "none") +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Distribution of the Mean by Sample Size")

path_fig <- projr::projr_path_get(
  "cache", "fig", "w6", "p-mean_example.pdf"
)
ggsave(
  path_fig, plot = p, width = 16, height = 14,
  units = "cm"
)
```

\centering
\begin{centering}
\includegraphics[width=0.575\textwidth]{`r path_fig`}
\end{centering}

# Problematic for confidence intervals - all problems fixed?

Suppose that we wish to estimate a confidence interval for the mean of some distritution.
Then, assuming that the data is normally distributed, we can use the following formula:

$$
\bar{x} \pm z_{\alpha/2} \hat{se}(\bar{x})
$$

where:

- $\bar{x}$ is the sample mean,
- $s$ is the sample standard deviation,
- $n$ is the sample size, and
- $z_{\alpha/2}$ is the critical value for the desired confidence level.

\pause

- Problem? $s$ is estimated

# We can improve this using the $t$-distribution

$$
\bar{x} \pm t_{\alpha/2, n-1} \hat{se}(\bar{x})
$$

where $t_{\alpha/2, n-1}$ is the critical value for the $t$-distribution with $n-1$ degrees of freedom.

- Further problems?
  - Assumes a symmetric sampling distribution
  - Assumes that $s$ and $\bar{x}$ are independent

# Bootstrapping

- **Definition**
  - A resampling technique used to perform inference sampling a dataset with replacement.

# Sampling with replacement

```{r}
orig_vec <- rnorm(1:3) |> signif(2)
boot_tbl <- tibble::tibble(
  Original = orig_vec,
  `Sample 1` = sample(orig_vec, replace = TRUE),
  `Sample 2` = sample(orig_vec, replace = TRUE),
  `Sample 3` = sample(orig_vec, replace = TRUE)
)
boot_tbl
```

- Mean for each boostrap sample: `r sapply(boot_tbl[, -1], mean) |> signif(2)`

# "Ideal" way to estimate a sampling distribution

- Repeatedly resample from the actual data generating process!

\begin{figure}[H]
\centering
\includegraphics[width=0.7\textwidth]{_raw_data/img/ideal_world.png}
\caption{Hesterberg (2014)}
\end{figure}

# Practical way to estimate a sampling distribution

- Repeatedly resample from the *sample*

\begin{figure}[H]
\centering
\includegraphics[width=0.7\textwidth]{_raw_data/img/bootstrap_world.png}
\caption{Hesterberg (2014)}
\end{figure}

# Bootstrap alternatives to $t$ confidence intervals I: The percentile method

1. Generate $B$ bootstrap samples.
2. Calculate the sample mean for each sample.
3. Set the (1-$\alpha$) confidence interval as the $\alpha/2$ and $1-\alpha/2$ percentiles of the bootstrap means.

# Implementation in R

- First, create our bootstrap sample:

```{r}
set.seed(3)
sample_vec <- rexp(10, rate = 0.1)^1.5
mean_vec_perc <- replicate(2e3, {
  sample(sample_vec, replace = TRUE) |> mean()
})
```

- Then, calculate our bounds:

```{r}
quantile(mean_vec_perc, c(0.025, 0.975)) |> signif(2)
```

```{r}
#| include: false
ci_vec_perc <- quantile(mean_vec_perc, c(0.025, 0.975))
```

# Bootstrap alternatives to $t$ confidence intervals I: Estimate se using the bootstrap

This has the following formula:

$$
\bar{x} \pm t_{\alpha/2, n-1} \hat{se^{\ast}}(\bar{x})
$$
  
  where $\hat{se^{\ast}}(\bar{x})$ is the standard deviation of the bootstrap means:

$$
\hat{se^{\ast}}(\bar{x}) = \sqrt{\frac{1}{B-1} \sum_{b=1}^{B} (\bar{x}^{(b)} - \bar{x})^2}
$$

# Bootstrap alternatives to $t$ confidence intervals II: Estimate se using the bootstrap

1. Generate $B$ bootstrap samples.
2. Calculate the sample mean for each sample.
3. Calculate the standard deviation of the bootstrap means.
4. Set the (1-$\alpha$) confidence interval as $\bar{x} \pm z_{\alpha/2} \hat{se}^{\ast}(\bar{x})$

# Implementation in R

```{r}
set.seed(3)
sample_vec <- rexp(10, rate = 0.1)^1.5
mean_actual <- mean(sample_vec)
mean_vec_perc <- replicate(2e3, {
  sample(sample_vec, replace = TRUE) |> mean()
})
se_boot <- sd(mean_vec_perc)
ci_vec_se_boot <- mean_actual +
  c(-1, 1) * se_boot * 
  qt(0.975, df = length(mean_vec_perc) - 1)
ci_vec_se_boot |> signif(2)
```

# Problem?

- The t-stat is not $t$-distributed!
- By $t$-test, we mean:

$$
\frac{\hat{\theta} - \theta}{\hat{se}(\hat{\theta})}
$$

- Why not?
  - Observations not normally distributed
  - Sample mean and sample standard deviation are not independent

# Sample standard deviation and mean correlated

```{r}
set.seed(3)
sample_vec <- rexp(10, rate = 0.1)^1.5
summ_stat_list <- replicate(2e3, {
  boot_sample <- sample(sample_vec, replace = TRUE)
  c("mean" = mean(boot_sample), "sd" = sd(boot_sample))
})
```

```{r}
#| include: false
plot_tbl <- tibble(
  mean = summ_stat_list[1, ],
  sd = summ_stat_list[2, ]
)
p <- ggplot(
  plot_tbl, aes(x = mean, y = sd)
) +
  geom_point(alpha = 0.25, colour = "dodgerblue") +
  labs(x = "Mean", y = "Standard Deviation") +
  cowplot::background_grid(major = "xy", minor = "none")
path_fig <- projr::projr_path_get(
  "cache", "fig", "w6", "p-mean_sd_corr.pdf"
)
ggsave(
  path_fig, plot = p, width = 10, height = 7,
  units = "cm"
)
```

\centering
\begin{centering}
\includegraphics[width=0.5\textwidth]{`r path_fig`}
\end{centering}

# Bootstrap alternatives to $t$ confidence intervals III: bootstrap the t-statistic

1. Generate $B$ bootstrap samples.
2. For each bootstrap sample, calculate the t-statistic.
3. Calcalate the $\alpha/2$ and $1-\alpha/2$ percentiles of the t-statistics.
3. Set the (1-$\alpha$) confidence interval as 

$$
\hat{\theta} - q_{1-\alpha/2}\hat{se}(\hat{\theta}), \hat{\theta} - q_{\alpha/2}\hat{se}(\hat{\theta})
$$

# Implementation in R {.smaller}

```{r}
set.seed(3)
sample_vec <- rexp(10, rate = 0.1)^1.5
mean_actual <- mean(sample_vec)
sd_actual <- sd(sample_vec) / sqrt(length(sample_vec) - 1)
t_stat_vec <- replicate(2e3, {
  boot_sample <- sample(sample_vec, replace = TRUE)
  mean_boot <- mean(boot_sample)
  se_boot <- sd(boot_sample) / sqrt(length(boot_sample) - 1)
  (mean_boot - mean_actual) / se_boot
})
quant_vec_t <- quantile(t_stat_vec, c(0.025, 0.975))
lb <- mean_actual - quant_vec_t[2] * sd_actual
ub <- mean_actual - quant_vec_t[1] * sd_actual
c(lb, ub) |> signif(2)
```

```{r}
#| include: false
ci_vec_boot_t <- c(lb, ub)
```

# Comparison of confidence intervals

```{r}
#| include: false
set.seed(3)
sample_vec <- rexp(10, rate = 0.1)^1.5
mean_actual <- mean(sample_vec)
sd_actual <- sd(sample_vec) / sqrt(length(sample_vec) - 1)
ci_vec_t <- mean_actual +
  c(-1, 1) * qt(0.975, df = length(sample_vec) - 1) * sd_actual
```

```{r}
#| echo: false
tibble(
  Method = c("T", "Percentile", "Bootstrap SE", "Bootstrap t"),
  `Lower bound` = c(ci_vec_t[1], ci_vec_perc[1], ci_vec_se_boot[1], ci_vec_boot_t[1]),
  `Upper bound` = c(ci_vec_t[2], ci_vec_perc[2], ci_vec_se_boot[2], ci_vec_boot_t[2])
)
```

# One-side non-coverage for normally distributed data

\begin{figure}[H]
\centering
\includegraphics[width=0.55\textwidth]{_raw_data/img/normal_sim.png}
\caption{Hesterberg (2014)}
\end{figure}

# One-side non-coverage for exponentially-distributed data

\begin{figure}[H]
\centering
\includegraphics[width=0.55\textwidth]{_raw_data/img/exp_sim.png}
\caption{Hesterberg (2014)}
\end{figure}

# Take-home messages

- Bootstrapping is a resampling technique for improved inference.
- The choice of confidence interval method dramatically affects coverage probabilities:
  - Small sample sizes: the percentile and bootstrap-se methods are too narrow. The percentile method performs atrociously at small sample sizes.
  - Skew data: the $t$- and bootstrap-se methods are symmetric, and the percentile method is insufficiently skew.
  - The bootstrap-$t$ method has good small sample performance and is more robust to skewness.
- Estimating confidence intervals for skew data is hard, and takes more samples than typically thought.

# References