---
title: "Regression Week I: Introduction and foundations to GLMs"
subtitle: "Lecture IV: The normal distribution and related theorems"
author: "Yovna Junglee (edited by Miguel Rodo)"
institute: University of Cape Town
aspectratio: 169
theme: Boadilla
format:
  beamer:
    embed-resources: true
    colorlinks: true
    include-in-header:
      file: preamble.tex
    classoption: "aspectratio=169"
---

# The centrality of the normal distribution

- As mentioned in the demonstration linear regression analysis, we typically assume that the errors are normally distributed.

<!-- \pause !-->

- This assumption leads to our parameters of interest (i.e. $\beta_0$, $\beta_1$, $\beta_2$, $\ldots$) having a normal distribution (which we'll show next week). 
- Therefore both our criterion for estimating the parameters as well as our inference regarding the parameters strongly depend on the normal distribution:

```{r, include = FALSE}
library(tibble)
library(dplyr)

data_raw_ad <- read.csv("_data_raw/Advertising.csv", header = TRUE)

# Convert the raw data to a tibble for easier manipulation
data_raw_ad <- as_tibble(data_raw_ad)

data_tidy_ad <- data_raw_ad |>
  dplyr::select(-X)
data_tidy_ad[1, ]

fit_tv <- lm(sales ~ TV, data = data_tidy_ad)
```

```{r , eval = FALSE, echo = FALSE}
summary(fit_tv)
```

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{_data_raw/img/lm_output.png}
\end{figure}


# Lecture outline

- Describe the PDFs and MGFs of the univariate and multivariate normal distributions
- Prove several results regarding the multivariate normal (MVN) distributions:
  - Linear combination of a MVN random variable is MVN
  - MVN composed of MVNs
  - Relationship between MVN and a $\chi^2$ random variable

# Exclusions compared to to previous years' courses

- The following appear in previous years' slides and tests (and in some cases, the notes), but won't be examined here:
  - Deriving MVN distribution
  - Deriving MVN MGF
  - Determining mean and variance of an MVN distribution given the exponent

<!-- \pause !-->

- The reasons are as follows:
  - Later results complete without them
  - Learning them well does not particularly help with other course content
  - Overlap with STA2004F
  - Adds significantly heavier calculus component

<!-- \pause !-->

- Note that being apply to make use of the MGF is still important.
- In general, however, the foundational mathematical facts and ideas of the coure are in linear algebra and the statistical distributions connected to the normal.


# The univariate normal distribution

::::{.columns}

:::{.column width="50%"}

\vspace{1.15cm}

::: {.def-univ_normal_dbn}

## Definition

If $X \sim \mathcal{N}(\mu,\sigma^2)$, then the probability density function is given as:

$$f(x)= \frac{1}{\sqrt{2 \pi \sigma^2}}\exp\left( -\frac{(x-\mu)^2}{2 \sigma^2} \right)$$
where $E(X)= \mu$ and $\textrm{Var}(X)=\sigma^2$ for $x\in\mathbb{R}$.
:::

:::

:::{.column width="50%"}

\vspace{0cm}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{_data_raw/img/rnorm_0_1.pdf}
    \caption{$X \sim \mathcal{N}(0,1)$}
    \label{fig:sims_rnorm}
\end{figure}


:::

::::