---
title: "Regression Week I: Introduction and foundations to GLMs"
subtitle: "Slide deck I: Course introduction and motivation"
author: Miguel Rodo
institute: University of Cape Town
date: 22 July 2024
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: preamble.tex
---

```{r}
#| include: false
library(ggplot2)
```

# What is the purpose of the regression section?

<!-- \pause !-->

- To make you a useful statistician!

<!-- \pause !-->

- Well, to really get you on your way
  - First year (STA1006S) is useful, but easily replicable by non-statisticans
  - STA2004F is (despite all the effort you put in) not immediately of great practical use
  - The practical skills and mathematical understanding you'll gain here are not easily replicable and can actually be used to assist decision-making

<!-- \pause !-->

## Key outcomes

- Mathematical understanding of simplest regression model
- Ability to code up an end-to-end modelling analysis
  - Reading in data
  - Analysing data
  - Displaying and interpreting results

# Lecture schedule

```{r }
#| echo: false
#| results: asis
#| label: tab-schedule
schedule_tbl <- tibble::tibble(
  Weeks = c("1", "2-6", "Vac", "7-11", "12"),
  `Start date` = c("22 Jul", "29 Jul", "30 Jul", "9 Sep", "14 Oct"),
  `End date` = c("26 Jul", "30 Aug", "8 Sep", "11 Oct", "18 Oct"),
  Section = c("Regression", "Experimental design", "")[
    c(1, 2, 3, 1, 2)
  ]
)
```

```{r}
#| echo: false
#| results: asis
#| label: tab-schedule-latex
#| eval: true
# Create and display the table
knitr::kable(schedule_tbl, format = "latex", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
```

# Contributions to regression section mark

```{r}
#| include: false
# Create the data frame
plot_tbl <- data.frame(
  pos = 1:3 |> rev(),
  percentage = c(70, 15, 15) |> rev(),
  category = 1:3 |> rev() |> as.character()
)

plot_tbl_txt <- tibble::tibble(
  x = 1:3,
  y = c(7.5, 7.5, 35) |> rev(),
  txt = c("70%", "15%", "15%")
)

plot_tbl_date <- tibble::tibble(
  x = 2:3,
  y = c(17, 17),
  txt = c("10 October", "30 September")
)

cat_lab_vec <- c("Assignment", "Test", "Exam") |> rev()

# Create the plot
p <- ggplot(plot_tbl, aes(x = pos, y = percentage, fill = category)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_x_continuous(
    labels = cat_lab_vec,
    breaks = 1:3
  ) +
  # geom_hline(yintercept = c(15, 70), linetype = "dashed") +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1) # ,
    # breaks = seq(0, 100, 10)
  ) +
  cowplot::theme_cowplot() +
  theme(
     axis.title.y = element_blank(),
     axis.title.x = element_blank(),
    # axis.ticks.y = element_blank(),
    # axis.text.y = element_blank(),
    legend.position = "none"
  ) +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) + #+
  scale_fill_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb")) +
  geom_text(
    data = plot_tbl_txt, aes(x = x, y = y, label = txt),
    size = 5, inherit.aes = FALSE
    ) +
  geom_text(
    data = plot_tbl_date, aes(x = x, y = y, label = txt),
    size = 5, inherit.aes = FALSE, hjust = 0
    )

path_p <- projr::projr_path_get(
  "cache", "fig", "w1s1", "p-assessment.png"
)
cowplot::ggsave2(
  path_p, p,
  width = 15, height = 8, units = "cm"
)
```

```{r}
#| results: asis
#| label: p-assessment
knitr::include_graphics(path_p)
```


# Learning resources: content

- Regression notes:
  - Location: `Module 1: Regression/More regression resources/Notes/NotesReg.pdf`
  - Printed copies available at stats reception later this week
- Slides:
  - Location: in `Slides` folder under each week's tab
- Lecture recordings:
  - Automatically uploaded to `Lecture Videos/Lecture Videos 2024` tab

# Learning resources: testing I

::::{.columns}

:::{.column width=50%}

- Q&A docs:
  - Comprehensive set of short questions on each slide deck's content
  - A review of all the facts, essentially
  - Will be in the `Slides` folder for each week's content
    - For example, `QARegSlidesW0S1PrereqLA.pdf` is in `Introduction/Slides` folder alongside `SlidesRegW0S1PrereqLA.pdf`
  - *Note*: these are not exam-style questions, but rather a way to a) remember the content, and b) check your understanding


:::


\pause

:::{.column width=50%}


- Tutorials
  - On Amathuba (`Module 1: Regression/More regression resources/Tutorials`)

\pause

- Question bank:
  - Appropriate questions over and above the tutorials
  - Aim to progress from simpler to harder questions within each section
  - On Amathuba (`Module 1: Regression/More regression resources/Question bank`)
  - Will be updated throughout the term

:::

::::

# Advised learning/revision route for exam and test

- Attend lectures
- After each lecture, do relevant Q&A content
- Do tutorials (and attend them)
- Do question bank questions

# Introduction to `R`

- Online lecture:
  - Day: Wednesday 24 July
  - Time: 2pm
  - Link: [MS Teams](https://teams.microsoft.com/l/meetup-join/19%3ameeting_NjhjYTBlYjgtNDk3Ni00MGUxLWEyYmQtZGM4YTM5NjljNjUy%40thread.v2/0?context=%7b%22Tid%22%3a%2292454335-564e-4ccf-b0b0-24445b8c03f7%22%2c%22Oid%22%3a%22f479858a-bdb3-4e57-9ee4-f624b487e11f%22%7d)

\pause

- Practicals:
  - Days: Thursday and Friday, 25 and 26 July
  - Times: 2-3pm and 3-4pm
  - Category: `Week 1 Computer Lab Session - Intro to R`

# Practicals

- In a computer lab, with a tutor, who will walk through coding up part of a regression analysis
- For example:
  - Fitting a regression model
  - Validating a regression model

# Sign up for tutorials and practicals

- Frequency: weekly
- Start: week 2
- Signup:
  - `Amathuba` >> `Course info` >> `Groups`
  - Categories;
    - Practical: `Computer Lab Session` and `STA2005SLabSession 5`
    - Tutorial: `STA2005S Tutorials`

# Office hours

- Room: 5.52 PD Hahn (fifth floor, down the corridor from stats reception)
- Time: 10-12
- Day: Thursday
- Purpose: any content-related questions
- May also email for other times, if need be

\pause

- Email:
  - Address: `miguel.rodo@uct.ac.za`
  - Purpose: any admin-related issues


# What is regression? {.smaller}

::::{.columns}

:::{.column width=50%}

- $Y\sim f(X,\epsilon)$:
  - $Y$ is the response variable (dependent variable)
  - $X$ contains the explanatory variable(s) (or independent variables) 
  - $\epsilon$ is the random variable
  - The function `f` describes how they relate
  - $Y$ cannot be categorical

:::

:::{.column width=50%}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{_data_raw/img/reg_img.png}
\end{figure}
:::

::::

# Regression versus the $t$-test

- Suppose you want to understand the relationship between blood pressure (the dependent variable) and taking a drug (the independent variable).
- Also suppose that more males took the drug than females.
- Suppose we perform a $t$-test. How might someone object?

<!--
- Confounders
!-->

\pause

- Regression allows us to control for confounders

# Regression versus the correlation coefficient

- Now, suppose a doctor wants to decide whether a six-month old baby has a healthy weight.
- Why would the correlation coefficient not be useful?

\pause

- Regression allows us to predict

# Toolbox

\centering

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{_data_raw/img/toolbox-wide-edit-removebg.png}
\end{figure}

# Assumptions

::::{.columns}
:::{.column width=70%}
```{r}
# ==========================
# distribution
# ==========================
# density
set.seed(3)
# generate data from a normal distrbution from a range of x values
p_normal <- ggplot(
  tibble::tibble(x = rnorm(1000, 200, 5)),
  aes(x = x)
) +
  geom_density(fill = "#a6cee3", adjust = 3, col = "gray65") +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
p_normal_non <- ggplot(
  tibble::tibble(x = rpois(1000, 200)^6),
  aes(x = x)
) +
  geom_density(fill = "#b2df8a", adjust = 3, col = "gray65") +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )

# ==========================
# relationship
# ==========================

set.seed(5)
n_elem <- 1e2
plot_tbl <- tibble::tibble(
  x = c(1, runif(n_elem, 1, 10), 10)
) |>
  dplyr::mutate(
    y = 2 + 3 * x + rnorm(n_elem + 2, 0, 2),
    y2 = 2 + 3 * x^3 + rnorm(n_elem + 2, 0, 300),
    fit = 2 + 3 * x,
    fit2 = 2 + 3 * x^3
  )

p_rel_linear <- ggplot(plot_tbl, aes(x = x, y = y)) +
  geom_point(col = "#a6cee3") +
  geom_line(aes(y = fit), color = "gray15", size = 1) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
p_rel_quad <- ggplot(plot_tbl, aes(x = x, y = y2)) +
  geom_point(col = "#b2df8a") +
  geom_line(aes(y = fit2), color = "gray15", size = 1) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )

# ==========================
# variance
# ==========================

set.seed(6)
n_elem <- 3e2
plot_tbl <- tibble::tibble(
  x = c(1, runif(n_elem, 1, 10), 10)
) |>
  dplyr::mutate(
    y = 1 * x + rnorm(n_elem + 2, 0, 2),
    y2 = (1 * x + rpois(n_elem + 2, x)^2)^0.9,
    y2 = y2 + rnorm(n_elem + 2, 0, 2)
  )

p_var_constant <- ggplot(plot_tbl, aes(x = x, y = y)) +
  geom_point(col = "#a6cee3") +
  geom_smooth(
    method = "lm", se = FALSE, col = "gray15", size = 1,
    formula = y ~ x
  ) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
p_var_increasing <- ggplot(plot_tbl, aes(x = x, y = y2)) +
  geom_point(col = "#b2df8a") +
 geom_smooth(
    method = "lm", se = FALSE, col = "gray15", size = 1,
    formula = y ~ x
  ) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )


p_grid <- cowplot::plot_grid(
  p_normal, p_rel_linear, p_var_constant,
p_normal_non, p_rel_quad, p_var_increasing,
  ncol = 3,
  labels = c("Distribution", "Relationship", "Variance"),
  label_size = 18
)
path_p_grid <- projr::projr_path_get(
  "cache", "fig", "w1s1", "p-assumptions.png"
)
cowplot::ggsave2(
  path_p_grid, p_grid,
  width = 19, height = 14, units = "cm"
)
```

```{r}
#| results: asis
knitr::include_graphics(path_p_grid)
```
:::

:::{.column width=30%}

\vskip 1cm

- In our course, we consider **linear regression** only:
  - Normally distributed response
  - Linear relationship with response variables
  - Constant variance

:::

::::

# Regression becomes highly flexible

- Non-constant variance: mean-variance relationship, dependent on predictor variables, etc.
- Correlated errors:
  - within subsets of observations, e.g. measurements from the same individual, or
  - between "nearby" observations, e.g. time-series or spatial data
- Non-normally distributed data, e.g. count data
- Categorial dependent vairables, e.g. disease status

# Mathematical foundations to linear regression

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{_data_raw/img/iceberg.png}
\end{figure}

# Developing inferential theory

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{_data_raw/img/lm_output.png}
\end{figure}

# Linear algebra and statistical theory prerequisites

- Under `Module 1: Regression/Introduction/Prerequisites`:
  - `NotesRegW0StatDbn.pdf`
    - Facts about normal, $t$ and $F$ distributions
  - `SlidesRegW0S1PrereqLA.pdf`
    - Essential linear algebra 
  - `QARegSlidesW0S1PrereqLA.pdf`
    - Questions and answers on the linear algebra content
  - Pre-recorded lecture video
    - Will upload one in the next day or two
  - `DemRegW0LinearAlgebra.html`
    - Implementing linear algebra in `R`
  - `TutReg1IntroMVNQuestions.pdf`
    - The first few questions in this tutorial are about linear algebra
  

