---
title: "Regression Week IV: ANOVA and Model Validation"
subtitle: "ANOVA"
author: "Miguel Rodo (with credit to Yovna Junglee)"
institute: "Department of Statistical Sciences - University of Cape Town"
bibliography: zotero.bib
format: 
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: preamble.tex
---



```{r}
#| include: false
library(tibble)
library(ggplot2)
theme_cowplot_bg <- function(font_size = 14) {
  cowplot::theme_cowplot(font_size = font_size) +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )
}
knitr::opts_chunk$set(echo = TRUE)
```

# Where we are in the course

- **Previous Topics Covered:**
  - Estimation of regression coefficients
  - Construction of confidence intervals
  - Hypothesis testing for linear combinations of coefficients
  - Simultaneous hypothesis tests
- **Next Steps:**
  - Describing ANOVA in the context of linear regression

# ANOVA: Analysis of Variance

- ANOVA is a statistical technique initially primarily used to compare the means of two or more groups.
- **Historical Context:**
  - Developed by Ronald Fisher in the early 20th century.
  - Served as a foundational method before the full development of inference for   general linear models.
- **Directly Considers Sums of Squares:**
  - ANOVA focuses on partitioning total variability into components.
  - Emphasizes understanding the sources of variation in data.

# Note re running code

- Run the following code chunk to load the necessary libraries and functions, if you want to run the code in the slides:

```{r}
#| eval: false
library(tibble)
library(ggplot2)
theme_cowplot_bg <- function(font_size = 14) {
  cowplot::theme_cowplot(font_size = font_size) +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )
}
```

# One-way ANOVA example

- Suppose individuals are given a placebo or one of two treatments to lower their blood pressure:

```{r}
set.seed(1811)
n <- 10
placebo <- rnorm(n, mean = 120, sd = 10)
trt_a <- rnorm(n, mean = 110, sd = 10)
trt_b <- rnorm(n, mean = 120, sd = 10)
trt_tbl <- tibble(
  group = rep(c("Placebo", "Treatment A", "Treatment B"), each = n),
  bp = c(placebo, trt_a, trt_b)
)
```

# One-way ANOVA example

- The data looks like this:

```{r}
trt_tbl[c(1, 11, 21), ]
```

# Blood pressure by treatment category

```{r}
#| echo: false
drug_col_vec <- stats::setNames(
  c("#b2df8a", "#a6cee3", "#1f78b4"),
  c("Placebo", "Treatment A", "Treatment B")
)
p <- ggplot(
  trt_tbl |>
    dplyr::mutate(
      group = factor(group, levels = c("Placebo", "Treatment A", "Treatment B"))
    ),
  aes(x = group, y = bp, fill = group)
) +
  theme_cowplot_bg() +
  geom_boxplot() +
  cowplot::background_grid(major = "xy") +
  ggforce::geom_sina(alpha = 0.75) +
  labs(
    x = "Treatment Group",
    y = "Blood Pressure"
  ) +
  theme(
    legend.position = "none"
  ) +
  scale_fill_manual(
    values = drug_col_vec
  )
path_fig <- "_tmp/fig/w4/s1anova/simple_anova_example-boxplot.pdf"
if (!dir.exists(dirname(path_fig))) {
  dir.create(dirname(path_fig), recursive = TRUE)
}
cowplot::ggsave2(
  path_fig,
  p,
  width = 9,
  height = 4.5,
  dpi = 300
)
```

\begin{center}
\includegraphics[width=\textwidth]{`r path_fig`}
\end{center}

# One-way ANOVA model specification

Here is the model for a one-way ANOVA with $a$ groups.

- *ANOVA Model:*

  $$
  Y_{ij} = \mu + \alpha_i + e_{ij}
  $$

  - $\mu$ is the overall mean.
  - $\alpha_i$ is the effect of group $i$.
  - $e_{ij}$ is the random error term for the $j$-th observation from the $i$-th group.

- *Assumptions:*
  - The observations are normally distributed
    - Or, less strictly, the sample size is large
  - The observations are independent
  - Variance is equal across groups
  - $\sum \alpha_i = 0$

# Hypothesis testing using the one-way ANOVA model

- **Null Hypothesis:**
  - $H_0: \alpha_1 = \alpha_2 = \ldots = \alpha_a = 0$
  - In our example, $a = 3$.
  - There is no difference in means across groups.
- **Alternative Hypothesis:**
  - $H_1: \text{At least one } \alpha_i \neq 0$
  - In our example, $H_1: \text{At least one of } \alpha_1,\;\alpha_2 \text{ and } \alpha_3 \neq 0$
  - There is a difference in means across groups.

# Formulae for $SS_A$ and $SSE$

::::{.columns}

:::{.column width="50%"}

- **$SS_A$ (between-group sum of squares):**

  $$
  SS_A = \sum_{i=1}^{a} n_i (\bar{Y}_{i\cdot} - \bar{Y}_{\cdot\cdot})^2
  $$

  - $n_i$ is the number of observations in group $i$.
  - $\bar{Y}_{i\cdot}$ is the mean of group $i$.
  - $\bar{Y}_{\cdot\cdot}$ is the overall mean.

:::

:::{.column width="50%"}

- **$SSE$ (within-group sum of squares):**

  $$
  SSE = \sum_{i=1}^{a} \sum_{j=1}^{n_i} (Y_{ij} - \bar{Y}_{i\cdot})^2
  $$

:::

::::

# Test statistic for one-way ANOVA

- The test statistic for ANOVA is the $F$-statistic:

  $$
  F = \frac{SS_A / (a - 1)}{SSE / (n - a)} = \frac{MS_A}{MSE}
  $$

  - $SS_A$ is the sum of squares between groups.
  - $SSE$ is the sum of squares within groups.
  - $a$ is the number of groups.
  - $n$ is the total number of observations.
  - $MS_A$ is the mean square between groups.
  - $MSE$ is the mean square within groups.

- Under the null hypothesis, $F \sim F(a - 1, n - a)$.

# Calculating the above from first principles

```{r}
y_bar <- mean(trt_tbl$bp)
n_i <- table(trt_tbl$group)
y_j_bar <- tapply(trt_tbl$bp, trt_tbl$group, mean)
ssa <- sum(n_i * (y_j_bar - y_bar)^2)
sse <- sum(tapply(
  trt_tbl$bp, trt_tbl$group, function(x) sum((x - mean(x))^2)
  ))
ssa |> signif(3)
sse |> signif(3)
```

# Obtaining the p-value

We first calculate the F-statistic:

```{r}
a <- length(n_i)
n <- nrow(trt_tbl)
f_stat <- (ssa / (a - 1)) / (sse / (n - a))
f_stat |> signif(3)
```

We now calculate the p-value:

```{r}
1 - pf(f_stat, a - 1, n - a)
```

# Using the `anova` function

- The `oneway.test` function in R can be used to perform one-way ANOVA directly:

```{r}
oneway.test(bp ~ group, data = trt_tbl, var.equal = TRUE)
```

- ANOVA is so closely connected to linear models, however, that almost all functions performing it will either require a model object or fit a model "under the hood":

```{r}
fit <- lm(bp ~ group, data = trt_tbl)
anova(fit)
```

# Correspondence with regression

- In fact, standard output from `summary(lm(...))` includes an ANOVA test for the nullity of all non-intercept coefficients:

```{r}
summary(fit)
```

# ANOVA as a subset of linear regression

::::{.callout-tip appearance="minimal"}
"Analysis of variance is a special case of regression, and both are special cases of the
GLM ["(generalised linear model)"]. All ANOVA problems can be solved through regression, but the reverse is not
true: Not all regression problems can be solved using traditional ANOVA."
[@tabachnick07]
::::

- ANOVA is a special case of linear regression:
  - When all predictors are categorical variables represented by dummy variables, the linear regression model becomes equivalent to an ANOVA model.

# ANOVA as a subset of linear regression

- For this reason, the `anova` function accepts a linear model object as an argument:

```{r}
fit <- lm(bp ~ group, data = trt_tbl)
anova(fit)
```

- This yields the same results as the "direct" ANOVA test.

# Why bother describing ANOVA then?

- Statistical methodology is often described in terms of ANOVA.
  - Even though ANOVA is a subset of linear regression.
- Collaborators may be familiar with ANOVA rather than regression.
  - This is largely a result of their training:
    - ANOVA is essentially an hypothesis-testing approach, which by focusing on one statistic (the p-value) is easier to understand.
    - Regression adds in additional elements: estimation and prediction.
  - To work with them, we need to be able to communicate in their language.

# Handling multi-level categorical explanatory variables in regression

- One-way ANOVA, at this stage, is in fact beyond what we've recovered in regression thus far.
- To bring regression up to speed, we need to consider how to handle multi-level categorical variables.
- The key is to use dummy variables, but we need to be careful about how we do this.

# An approach that won't work: the model

- Imagine that we used the following model, where each level of a categorical variable got its own regression parameter:

$$
Y = \beta_0 + I(\text{group} = 1)\beta_1 + I(\text{group} = 2)\beta_2 + \ldots + I(\text{group} = a)\beta_a + e
$$

- Note that $I(\text{group} = i)$ is an indicator function that is 1 if the group is $i$ and 0 otherwise.
- For example, if an individual belongs to group 2, the model becomes:

$$
Y = \beta_0 + 0\beta_1 + 1\beta_2 + 0\beta_3 + \ldots + 0\beta_r + e  = \beta_0 + \beta_2 + e
$$

# An approach that won't work: the design matrix

- In this design matrix, we have $n$ rows and, importantly, $a+1$ columns: one for the intercept and one for each level of the categorical variable.
- Suppose that, for the sake of illustration, the first row of the design matrix belongs to category 1, the second row to category 2, and so on until the $a$-th row to category $a$:

$$
\symbf{X}_{1:r,:} = \begin{bmatrix}
1 & 1 & 0 & \ldots & 0 \\
1 & 0 & 1 & \ldots & 0 \\
\ldots & \ldots & \ldots & \ldots & \ldots \\
1 & 0 & 0 & \ldots & 1
\end{bmatrix}
$$

# An approach that won't work: the design matrix

- If $a=3$, the first two rows of the design matrix would be:

$$
\symbf{X}_{1:3,:} = \begin{bmatrix}
1 & 1 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 0 & 1
\end{bmatrix}
$$

- All the rows below repeat one of these first three rows.
- Clearly, therefore, the columns are linearly dependent, as the first column is the sum of the other columns.

# An approach that won't work: the estimator

- The linear dependence of the estimator makes $\symbf{X}'\symbf{X}$ non-invertible, and therefore

  $$
  \hat{\symbf{\beta}} = (\symbf{X}'\symbf{X})^{-1}\symbf{X}'\symbf{Y}
  $$

  does not exist.

# Adopting a reference level

- When each level of categorical variable got its own parameter, we would tend to think of $\beta_0$ as the overall mean, like $\mu$ in ANOVA.
- However, in ANOVA this is achieved by constraining the group effects to sum to zero ($\sum \alpha_i = 0$).
- In regression, we could do that, but typically we adopt a reference level.
- The reference level is the level for which $\beta_0$ is the mean (assuming all other explanatory variables are zero).

# Adopting a reference level

- We achieve a reference level by creating dummy variables for all but one level.
- If we drop the $a$-th level, the model then becomes

$$
Y = \beta_0 + I(\text{group} = 1)\beta_1 + I(\text{group} = 2)\beta_2 + \ldots + I(\text{group} = a-1)\beta_{a-1} + e
$$

- The mean of group $a$ is then $\beta_0$.
- The mean of group $i$ is $\beta_0 + \beta_i$.
- The interpretation of $\beta_i$ is therefore the difference in means between group $i$ and group $a$.
  - Hence the term "reference" level.

# The design matrix used by `lm` for our drug example

- Let's view the rows of the design matrix ($\symbf{X}$) of our `group` variable, corresponding to the three groups:

```{r}
model.matrix(fit)[c(1, 11, 21), ]
```

- The first row corresponds to the reference level (placebo).
- The second row corresponds to treatment A.
- The third row corresponds to treatment B.

# Translation of one-way ANOVA to the regression: the null hypothesis

- The null hypothesis in ANOVA is equivalent to the following null hypothesis in regression:

  $$
  H_0: \symbf{L}\symbf{\beta} = \symbf{L}\symbf{\beta}^{(0)} = \symbf{0}
  $$

- where:
  - $\symbf{L}$ is an $(a-1)\times k$ with one 1 per row and 0s otherwise, e.g. $\begin{bmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}$
  - The non-zero column indices in $\symbf{L}$ correspond to the positions of the categorical variable's regression parameters in $\symbf{\beta}$.
  - For example, in our drug example, $\symbf{L} = \begin{bmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}$ as $\symbf{\beta}' = \left(\beta_0, \beta_{A}, \beta_B\right)$
  - $\symbf{\beta}^{(0)} = \symbf{0}$

# Translation of one-way ANOVA to the regression: the null hypothesis

- Alternatively, 

  $$
  H_0: \beta_i = 0 \text{ for all } i \in \mathbb{A}
  $$

  for $\mathbb{A}$ the positions in $\symbf{\beta}$ corresponding to the categorical variable's effects.

# Translation of one-way ANOVA to regression: the $F$-test

- The one-way ANOVA $F$-statistic is equivalent to the following $F$-statistic in regression:

$$
F = \frac{(\symbf{L}\symbf{\beta} - \symbf{L}\symbf{\beta}^{(0)})'(s^2\symbf{L}\symbf{C}\symbf{L}^T)^{-1}(\symbf{L}\symbf{\beta} - \symbf{L}\symbf{\beta}^{(0)})}{a-1} = \frac{(\symbf{L}\symbf{\beta})'(s^2\symbf{L}\symbf{C}\symbf{L}^T)^{-1}(\symbf{L}\symbf{\beta})}{a-1} 
$$

- where:
  - $\symbf{L}$ is an $(a-1)\times k$ matrix of zeros and ones, e.g. $\begin{bmatrix} 0 & 1 & 0 \\ 0 & 0 & 1 \end{bmatrix}$,
  - implicitly, $\symbf{\beta}^{(0)} = \symbf{0}$
- Under the null, $F \sim F(a - 1, n - a)$.

# The $F$-test for our drug example

- We can calculate the $F$-statistic for our drug example:

```{r}
L <- matrix(c(0, 1, 0, 0, 0, 1), nrow = 2, byrow = TRUE)
C <- solve(t(model.matrix(fit)) %*% model.matrix(fit))
s2 <- summary(fit)$sigma^2
coef_mat <- matrix(coef(fit), ncol = 1)
f_stat <- (t(L %*% coef_mat) %*% solve(L %*% C %*% t(L)) %*% (L %*% coef_mat)) / s2 / nrow(L)
1 - pf(f_stat, nrow(L), nrow(model.matrix(fit)) - ncol(model.matrix(fit)))
```

# Reduced and full models

- The $F$-test for a subset of coefficients being zero can be calculated in an alternative fashion using two model fits.
- Suppose that we have a reduced model, $M_0$, and a full model, $M_1$:
  - Under $M_1$:
    - Suppose that $Y=\symbf{X}\symbf{\beta} + \symbf{\epsilon}$.
  - Under $M_0$:
    - Let $\symbf{X}^{(0)}$ be a subset of the columns of $\symbf{X}$ (after excluding the columns corresponding to coefficients that are zero), and $\symbf{\beta}^{(0)}$ be the corresponding estimates.
    - Then $Y=\symbf{X}^{(0)}\symbf{\beta}^{(0)} + e$.

# Sums of squares for the reduced and full models

- Let $\hat{\symbf{Y}}$ and $\hat{\symbf{Y}}^{(0)}$ be the predicted values under the full and reduced models, respectively.
- Let $SSE_{UR}$ be the sum of squared errors under the unrestricted (full) model and $SSE_{R}$ be the sum of squared errors under the restricted (reduced) model.
- Suppose that the number of parameters in the full and reduced models are $k_{UR}$ (just $k$, really) and $k_{R}$ (some value less than $k$), respectively.
- Then

  $$
  SSE_{UR} = \sum_{i=1}^{n} (Y_i - \hat{Y}_i)^2, \quad SSE_{R} = \sum_{i=1}^{n} (Y_i - \hat{Y}_i^{(0)})^2.
  $$

# The $F$-test using the reduced and full models

- The $F$-statistic is then

  $$
  F = \frac{(SSE_{R} - SSE_{UR}) / (k_{UR} - k_{R})}{SSE_{UR} / (n - k_{UR})}
  $$

- Under the null hypothesis, $F \sim F(k_{UR} - k_{R}, n - k_{UR})$.

# Application of this $F$-test calculation to our drug example

- We can calculate the $F$-statistic for the full and reduced models in our drug example:

```{r}
fit_reduced <- lm(bp ~ 1, data = trt_tbl)
fit_full <- lm(bp ~ group, data = trt_tbl)
sse_ur <- sum(residuals(fit_full)^2)
sse_r <- sum(residuals(fit_reduced)^2)
k_ur <- length(coef(fit_full))
k_r <- length(coef(fit_reduced))
f_stat <- ((sse_r - sse_ur) / (k_ur - k_r)) / (sse_ur / (n - k_ur))
1 - pf(f_stat, k_ur - k_r, nrow(trt_tbl) - k_ur)
```

# Using the `anova` function to compare two models

```{r}
anova(fit_reduced, fit_full)
```

# F-statistic in the output of `summary(lm(...))`

- The $F$-statistic in the output of `summary(lm(...))` tests the null hypothesis that all non-intercept coefficients are zero.
- In other words, it tests the following null hypothesis:

  $$
  H_0: \beta_1 = \beta_2 = \ldots = \beta_k = 0
  $$
- It is useful for testing whether the explanatory variablles provide any information at all.

# F-statistic in the output of `summary(lm(...))`

- As the $F$ tests above assessed whether either drugs were different from placebo, they are equivalent to the $F$-statistic in the output of `summary(lm(...))`:

```{r}
#| results: hide
summary(lm(bp ~ group, data = trt_tbl))
```

\begin{center}
\includegraphics[width=0.6\textwidth]{_raw_data/img/f_stat.png}
\end{center}

# Two-way ANOVA

- Two-way ANOVA is an extension of one-way ANOVA that allows for two categorical explanatory variables.
- Here is the model statement for two-way ANOVA:

$$
Y_{ijk} = \mu + \alpha_i + \beta_j + e_{ijk}
$$

- where:
  - $\mu$ is the overall mean.
  - $\alpha_i$ is the effect of the $i$-th level of the first categorical variable.
  - $\beta_j$ is the effect of the $j$-th level of the second categorical variable.
  - $e_{ijk}$ is the random error term for the $k$-th observation from the $i$-th level of the first variable and the $j$-th level of the second variable.

# Simulating our advertising data again

```{r}
# Advertising data
set.seed(201)
n_ad <- 100L
spend_vec_radio <- runif(n_ad, 0, 3)
spend_vec_tv <- 1.2 * spend_vec_radio + runif(n_ad, 0, 3)
revenue_vec <- 2 * spend_vec_tv + rnorm(n_ad, mean = 10, sd = 2)

ad_tbl <- tibble(
  spend_radio = spend_vec_radio,
  spend_tv = spend_vec_tv,
  revenue = revenue_vec
)
```

# Extracting p-values from simple and multiple linear regression models

```{r}
fit_simple_tv <- lm(revenue ~ spend_tv, data = ad_tbl)
fit_simple_radio <- lm(revenue ~ spend_radio, data = ad_tbl)
fit_multiple <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
```

```{r}
#| include: false
p_val_tbl_lm <- tibble(
  Model = c("Simple TV", "Simple Radio", "Multiple"),
  `P-value (TV)` = c(
    summary(fit_simple_tv)$coefficients[2, 4],
    NA,
    summary(fit_multiple)$coefficients[2, 4]
  ),
  `P-value (Radio)` = c(
    NA,
    summary(fit_simple_radio)$coefficients[2, 4],
    summary(fit_multiple)$coefficients[3, 4]
  )
)
```

```{r}
#| echo: false
p_val_tbl_lm
```

# Fitting a two-way ANOVA model

- The order of the explanatory variables matters for the "default" ANOVA test.

- First we fit a two-way ANOVA model with TV spend first:

```{r}
fit_tv_first <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
anova(fit_tv_first)
```

# Fitting a two-way ANOVA model

- Now we fit a two-way ANOVA model with radio spend first:

```{r}
fit_radio_first <- lm(revenue ~ spend_radio + spend_tv, data = ad_tbl)
anova(fit_radio_first)
```

# Final comparison of p-values

```{r}
#| echo: false
anova_radio_first <- anova(fit_radio_first)
p_anova_radio_first <- anova_radio_first$`Pr(>F)`[1:2]
anova_tv_first <- anova(fit_tv_first)
p_anova_tv_first <- anova_tv_first$`Pr(>F)`[1:2]
p_val_tbl_anova <- tibble(
  Model = c("ANOVA (TV first)", "ANOVA (Radio first)"),
  `P-value (TV)` = c(
    p_anova_tv_first[1],
    p_anova_radio_first[2]
  ),
  `P-value (Radio)` = c(
    p_anova_tv_first[2],
    p_anova_radio_first[1]
  )
)
p_val_tbl_I <- p_val_tbl_lm |>
  dplyr::bind_rows(p_val_tbl_anova)
p_val_tbl_I
```

- The key point is that multiple linear regression and ANOVA yield the same p-value for the *last* explanatory variable included in the ANOVA model.

# Type I vs Type II ANOVA

- The previous ANOVA example was an example of **Type I ANOVA**.
  - This is also known as "sequential" ANOVA.
  - In type I ANOVA, the effects of the variables are estimated in the order they are included in the model.
- In type II ANOVA, the effects of the variables are estimated in a similar fashion to regression:
  - The effects of the first variable are estimated after the effects of the second variable.
  - This means that the order of the variables does not matter.

# Applying Type II ANOVA: TV first

```{r}
car::Anova(fit_tv_first, type = "II")
```

# Applying Type II ANOVA: Radio first

```{r}
car::Anova(fit_radio_first, type = "II")
```

# Multiple linear regression summary

```{r}
summary(fit_tv_first)$coefficients |> signif(3)
```

# Regression treats every variable as if it were the last one

- This in fact gives rise to the interpretation of regression coefficients as "holding all other variables constant".
- They're not so much "constant" as "we have already accounted for them".
- The justification comes from us getting the same regression coefficients between the following two scenarios:
  - Multiple linear regression with all variables included.
  - Simple linear regression for each variable *after* removing any association between each variable and the other variables in the model.

# Equivalent algorithm for regression

- In more detail, for e.g. $X_1$ we:
  - Fit `lm(X_1 ~ X_2 + X_3 + \ldots + X_k)`.
  - Extract the residuals, $\hat{\mathbf{e}}^{(1)}=\mathbf{X}_1 - \hat{\mathbf{X}}_1$.
  - Model $\mathbf{Y} = \mathbf{1}\beta_0 + \hat{\mathbf{e}}^{(1)}$.
  - The resulting coefficient is the same as the coefficient for $X_1$ in the multiple linear regression model (i.e. `lm(Y ~ X_1 + X_2 + \ldots + X_k)`).

# Demonstration

First, let's fit a multiple linear regression model:

```{r}
lm(revenue ~ spend_tv + spend_radio, data = ad_tbl) |>
  summary() |>
  broom::tidy()
```

# Demonstration cont.

Now, let's fit a simple linear regression model for `spend_radio` after removing the association with `spend_tv`:
  
```{r}
fit_radio <- lm(spend_radio ~ spend_tv, data = ad_tbl)
residuals_radio <- resid(fit_radio)
lm(revenue ~ residuals_radio, data = ad_tbl) |>
  summary() |> coefficients() |> signif(3)
```

# Demonstration cont.

Finally, let's fit a simple linear regression model for `spend_tv` after removing the association with `spend_radio`:

```{r}
fit_tv <- lm(spend_tv ~ spend_radio, data = ad_tbl)
residuals_tv <- resid(fit_tv)
lm(revenue ~ residuals_tv, data = ad_tbl) |>
  summary() |> coefficients() |> signif(3)
```

# Take-aways

- ANOVA is a statistical technique used to compare the means of two or more groups.
  - It can handle multiple groups and multiple levels of categorical variables.
- ANOVA is a special case of linear regression.
  - The $F$-statistic in ANOVA is equivalent to the $F$-statistic in regression for a subset of coefficients being zero.
    - This equivalency always holds in one-way ANOVA, but becomes more complicated in multi-way ANOVA.
- Regression uses dummy variables to handle categorical variables.
  - The reference level is the level for which the intercept is the mean.
  - The effect of each level is the difference in means between that level and the reference level.
  - For a categorical variable with $a$ levels, $a-1$ dummy variables (and therefore $a-1$ regression coefficients) are needed.
- The $F$-test can be calculated in three different ways:
  - Directly from the sums of squares (i.e. using $SS_A$ and $SSE$).
  - Using the full and reduced models (i.e. using $SSE_{UR}$ and $SSE_{R}$).
  - Using the linear regression model (i.e. using $\symbf{L}$ and $\symbf{C}$).

# Works cited

